---
title: "Test_JD1"
author: "Jessie Davidson"
format: html
editor: visual
---

#### Environmental Factors Predicting Solitary Bee Distributions

This is a project that will test a variety of environmental variables to discover whether any of these predict the distribution of solitary bees.

We begin by downloading and installing libraries to R

```{r}
#| echo: false
library(dplyr)

```

We import data that was exported by the Biodiverse package for Gilbert Alacon-Cruz thesis.

```{r}
#| echo: false
bee_data <- read.csv('~/Github/bees_climate/data/BeeDataNoiNatSpecimensPerCell.csv')

```

#### Practice with mapview package:

Start by installing mapview package and then access it in the library.

```{r}
#| echo: false
install.packages('mapview')
```

```{r}
#| echo: false
library(mapview)
```

Use the mapview() function.

```{r}
bee_data_map <- mapview(bee_data, xcol = "X.Axis_0.", ycol = "X.Axis_1.", crs = 3310, grid = TRUE)

print(bee_data_map)
```

Changing colors in mapview function:

```{r}
bee_data_map_colors <- mapview(bee_data, xcol = "X.Axis_0.", ycol = "X.Axis_1.", zcol = 'X..Agapostemon.texanus..', crs = 3310, grid = TRUE)

print(bee_data_map_colors)
```

Using RColorBrewer:

```{r}
library(RColorBrewer)
```

```{r}
bee_data_map_RColorBrewer <- mapview(bee_data, xcol = "X.Axis_0.", ycol = "X.Axis_1.", zcol = 'X..Agapostemon.texanus..', col.regions = ('Set1'), crs = 3310, grid = TRUE) # not really important, but getting an errot on this, might do more research to figure it out

print(bee_data_map_RColorBrewer)

```
