---
title: "California Bee Diversity & Conservation Gaps"
author: "Gretchen LeBuhn"
date: "`r Sys.Date()`"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 4
    theme: sandstone
    code-fold: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
---

## Overview

This document overlays California bee species richness and endemism (CWE) data with the CA Nature Terrestrial 30×30 Conserved Areas layer to identify conservation gaps — areas of high bee diversity or endemism that fall outside GAP Status 1 & 2 protected lands.

Three gap maps are produced:

-   **Fig. 1** — Conservation gaps: high richness, low protection
-   **Fig. 2** — Conservation gaps: high endemism, low protection
-   **Fig. 3** — Dual-gap priority cells: high in both richness and endemism, unprotected

A sensitivity analysis examines how gap counts change when the minimum areal-overlap threshold required to classify a cell as "protected" is varied from 10% to 75%.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup

# tidyverse loads ggplot2, tidyr, stringr, purrr, dplyr, readr, and more.
# Additional packages are listed only where they add something tidyverse lacks.
library(tidyverse)   # core data wrangling and plotting
library(sf)          # simple features: read, transform, join spatial data
library(viridis)     # perceptually uniform, colorblind-safe colour palettes
library(patchwork)   # combine multiple ggplot objects into one figure
library(scales)      # format axis labels and legend text
library(glue)        # string interpolation for labels and messages
library(tigris)      # Census Bureau CA state boundary
library(kableExtra)  # styled HTML tables with column spanners and footnotes

# Helper functions for the validation chunk (defined here so they are
# available throughout the document, not buried in the validation chunk).
checks <- tibble(
  severity    = character(),
  id          = character(),
  description = character(),
  detail      = character()
)

flag <- function(severity, id, description, detail = "") {
  checks <<- bind_rows(
    checks,
    tibble(severity, id, description, detail = as.character(detail))
  )
}
pass <- function(id, description, detail = "") flag("PASS", id, description, detail)
warn <- function(id, description, detail = "") flag("WARN", id, description, detail)
fail <- function(id, description, detail = "") flag("FAIL", id, description, detail)

# All data files live in bees_climate/data/, one level above this script.
data_dir <- file.path("..", "data")
```

------------------------------------------------------------------------

## Load Data

The two loading chunks below are kept separate so each data source is clearly documented on its own. Both are reprojected to CA Albers (EPSG 3310, units in metres) immediately on load so that every subsequent spatial operation uses a consistent coordinate system.

### Bee diversity data

The processed bee data contains ecoregion assignments and grid-cell centroids in CA Albers (EPSG 3310). The script prefers the GeoPackage (which preserves geometry and CRS exactly) and falls back to the CSV if the GeoPackage is absent.

```{r}
#| label: load-bee-data

processed_dir <- file.path(data_dir, "processed")
gpkg_path     <- file.path(processed_dir, "eco.data.gpkg")
csv_path      <- file.path(processed_dir, "eco.data.csv")

if (file.exists(gpkg_path)) {

  bee.data <- st_read(gpkg_path, quiet = TRUE)
  message("Loaded bee data from GeoPackage: ", gpkg_path)

} else if (file.exists(csv_path)) {

  message("GeoPackage not found — loading from CSV: ", csv_path)

  # Column names in the raw CSV are positional index strings produced by the
  # upstream export script (e.g. "X_Axis_0_" = x-centroid). The named vector
  # below maps them to readable names and is easy to update if they change.
  col_renames <- c(
    x.axis   = "X_Axis_0_",
    y.axis   = "X_Axis_1_",
    cwe      = "X_ENDC_CWE_",
    richness = "X_ENDC_RICHNESS_"
  )

  bee.data <- read_csv(csv_path, show_col_types = FALSE) |>
    rename(all_of(col_renames)) |>
    select(x.axis, y.axis, cwe, richness,
           starts_with("bio"), starts_with("L")) |>
    drop_na(x.axis, y.axis) |>
    st_as_sf(coords = c("x.axis", "y.axis"), crs = 3310)

} else {
  stop(
    "Neither eco.data.gpkg nor eco.data.csv found in: ", processed_dir,
    "\nCheck that bees_climate/data/processed/ contains the bee data files."
  )
}

bee.data <- st_transform(bee.data, crs = 3310)

message(glue(
  "Bee data: {nrow(bee.data)} grid cells | ",
  "Richness range: {round(min(bee.data$richness, na.rm=TRUE), 1)} – ",
  "{round(max(bee.data$richness, na.rm=TRUE), 1)} | ",
  "CWE range: {round(min(bee.data$cwe, na.rm=TRUE), 3)} – ",
  "{round(max(bee.data$cwe, na.rm=TRUE), 3)}"
))
```

### Build 15 × 15 km square quadrats

Each point centroid is expanded into a square polygon with a 7,500 m half-side (= 15 km in EPSG 3310 metre units). The point geometry column in `bee.data` is replaced by these squares for all subsequent spatial operations.

```{r}
#| label: grid-to-squares

half_side <- 7500  # metres (half of 15 km)

# Returns a square sf::POLYGON centred on the input point geometry.
make_square_poly <- function(geom, half = half_side) {
  xy <- st_coordinates(geom)
  x  <- xy[1, "X"]
  y  <- xy[1, "Y"]
  st_polygon(list(matrix(
    c(x - half, y - half,
      x + half, y - half,
      x + half, y + half,
      x - half, y + half,
      x - half, y - half),   # close the ring
    ncol = 2, byrow = TRUE
  )))
}

square_geoms <- st_geometry(bee.data) |>
  lapply(make_square_poly) |>
  st_sfc(crs = 3310)

bee.data <- bee.data |> st_set_geometry(square_geoms)

message(glue(
  "Quadrats built: {nrow(bee.data)} squares | ",
  "geometry type: {unique(st_geometry_type(bee.data))}"
))
```

### 30×30 Conserved Areas shapefile

The CA Nature Terrestrial 30×30 layer is distributed as a zip archive. The chunk below locates the zip, extracts it to a temporary directory, and reads the shapefile into R.

```{r}
#| label: load-conserved

conserved_zip_name <- "30x30_Conserved_Areas, Terrestrial_(2025).zip"
zip_path           <- file.path(data_dir, conserved_zip_name)

# Also try the URL-encoded variant (some systems save commas as %2C).
if (!file.exists(zip_path)) {
  zip_path_encoded <- file.path(
    data_dir,
    "30x30_Conserved_Areas%2C_Terrestrial_(2025).zip"
  )
  if (file.exists(zip_path_encoded)) {
    zip_path <- zip_path_encoded
  } else {
    stop(
      "Conserved areas zip not found.\n",
      "Expected at: ", file.path(data_dir, conserved_zip_name), "\n",
      "Check that bees_climate/data/ contains the 30x30 zip file."
    )
  }
}

conserved_dir <- tempfile("conserved_")
dir.create(conserved_dir)
unzip(zip_path, exdir = conserved_dir)

conserved_path <- list.files(
  conserved_dir,
  pattern     = "\\.shp$",
  full.names  = TRUE,
  recursive   = TRUE,
  ignore.case = TRUE
)

stopifnot(
  "No .shp file found after unzipping — check the zip contents" =
    length(conserved_path) >= 1
)

# If the zip contains multiple shapefiles, use the first and warn.
if (length(conserved_path) > 1) {
  warning("Multiple .shp files found; using: ", basename(conserved_path[1]))
}
conserved_path <- conserved_path[1]

conserved <- st_read(conserved_path, quiet = TRUE) |>
  st_transform(crs = 3310)

message(glue("Conserved areas: {nrow(conserved)} polygons loaded."))
```

------------------------------------------------------------------------

## Prepare Conserved Areas

GAP Status codes indicate the management intent of a protected area:

-   **GAP 1** — permanently protected, primary biodiversity focus
-   **GAP 2** — managed mostly for biodiversity; may allow some extractive use
-   **GAP 3** — multiple use; biodiversity not primary focus
-   **GAP 4** — no known biodiversity focus (e.g. working lands)

This analysis defines "well conserved" as GAP 1 or 2.

### Filter and repair geometries

```{r}
#| label: filter-repair-conserved

# Keep only polygons with any GAP 1 or GAP 2 acreage.
conserved_gap12 <- conserved |>
  filter(Gap1_acres > 0 | Gap2_acres > 0)

message(glue(
  "Invalid geometries before repair: {sum(!st_is_valid(conserved_gap12))}"
))

# Drop Z/M coordinates (elevation or measure values can confuse st_make_valid),
# then repair any remaining invalid geometries.
conserved_gap12 <- conserved_gap12 |>
  mutate(geometry = st_zm(geometry, drop = TRUE, what = "ZM")) |>
  mutate(geometry = st_make_valid(geometry))

message(glue(
  "Invalid geometries after repair: {sum(!st_is_valid(conserved_gap12))}"
))
```

### Dissolve to a single union polygon

Dissolving all GAP 1 & 2 parcels into one polygon before the spatial join is critical for performance — joining against a single geometry is orders of magnitude faster than joining against tens of thousands of individual parcels.

```{r}
#| label: dissolve-conserved
#| cache: true

message("Dissolving GAP 1 & 2 areas... (may take a moment)")

conserved_union <- conserved_gap12 |>
  summarise(geometry = st_union(geometry)) |>
  mutate(protected = "GAP 1 & 2")

gap12_acres <- sum(
  conserved_gap12$Gap1_acres + conserved_gap12$Gap2_acres,
  na.rm = TRUE
)
message(glue("GAP 1 & 2 coverage: {round(gap12_acres / 1e6, 2)} million acres"))
```

------------------------------------------------------------------------

## Classify Bee Cells by Protection Status

Each 15 × 15 km bee grid cell is tested for intersection with the dissolved GAP 1 & 2 polygon. A left join retains all cells; those with no overlap receive `is_protected = FALSE`. This single `bee_prot` object is the source of truth for all downstream gap classification and sensitivity analyses.

> **Note on deduplication:** After dissolving `conserved_union`, a cell should match at most one polygon. The `group_by + slice(1)` below guards against edge-case duplicates from floating-point boundary touches — it does not silently drop real data.

```{r}
#| label: spatial-join
#| cache: true

bee_prot <- bee.data |>
  mutate(bee_cell_id = row_number()) |>
  st_join(conserved_union["protected"], join = st_intersects, left = TRUE) |>
  group_by(bee_cell_id) |>
  slice(1) |>
  ungroup() |>
  mutate(is_protected = !is.na(protected))

n_prot   <- sum(bee_prot$is_protected)
n_unprot <- sum(!bee_prot$is_protected)

message(glue(
  "Protection status: {n_prot} cells overlap GAP 1 & 2 ",
  "({round(100 * n_prot / nrow(bee_prot), 1)}%), ",
  "{n_unprot} cells unprotected ",
  "({round(100 * n_unprot / nrow(bee_prot), 1)}%)"
))
```

------------------------------------------------------------------------

## Define Conservation Gap Thresholds

"High" richness and endemism are cells at or above the 75th percentile of their respective distributions across all California grid cells. This threshold captures the top quarter of biodiversity values while retaining enough cells to be spatially meaningful. Adjust `gap_quantile` to change the definition without touching any other code.

```{r}
#| label: gap-thresholds

gap_quantile <- 0.75   # change here to tighten or loosen the "high" definition

rich_thresh <- quantile(bee.data$richness, gap_quantile, na.rm = TRUE)
cwe_thresh  <- quantile(bee.data$cwe,      gap_quantile, na.rm = TRUE)

message(glue(
  "Richness threshold (≥{gap_quantile*100}th pctile): {round(rich_thresh, 1)} species\n",
  "CWE threshold     (≥{gap_quantile*100}th pctile): {round(cwe_thresh, 4)}"
))

# Classify each cell into one of four richness × protection categories and
# one of four endemism × protection categories for map colouring.
bee_gaps <- bee_prot |>
  mutate(
    rich_gap = richness >= rich_thresh & !is_protected,
    cwe_gap  = cwe      >= cwe_thresh  & !is_protected,

    rich_priority = case_when(
      richness >= rich_thresh & !is_protected ~ "High richness, unprotected",
      richness >= rich_thresh &  is_protected ~ "High richness, protected",
      richness <  rich_thresh & !is_protected ~ "Lower richness, unprotected",
      TRUE                                    ~ "Lower richness, protected"
    ),

    cwe_priority = case_when(
      cwe >= cwe_thresh & !is_protected ~ "High endemism, unprotected",
      cwe >= cwe_thresh &  is_protected ~ "High endemism, protected",
      cwe <  cwe_thresh & !is_protected ~ "Lower endemism, unprotected",
      TRUE                              ~ "Lower endemism, protected"
    )
  )

dual_gap_n <- sum(bee_gaps$rich_gap & bee_gaps$cwe_gap)
message(glue(
  "Dual-gap cells (high richness AND endemism, unprotected): {dual_gap_n}"
))
```

------------------------------------------------------------------------

## Shared Map Elements

Three reusable objects are composed into every map panel:

-   **`map_theme`** — publication-ready theme (white background, bottom legend).
-   **`ca_outline_layer`** — CA state boundary from the Census Bureau via `tigris`, reprojected to CA Albers.
-   **`conserved_base`** — light grey-green fill showing GAP 1 & 2 parcels as context beneath all data layers.

All colours follow the Wong (2011) colorblind-safe palette and reproduce clearly in greyscale.

```{r}
#| label: map-theme

bg_col       <- "#ffffff"
text_col     <- "#1a1a1a"
subtitle_col <- "#555555"
border_col   <- "#333333"

map_theme <- theme_void(base_size = 11) +
  theme(
    plot.background  = element_rect(fill = bg_col, color = NA),
    panel.background = element_rect(fill = bg_col, color = NA),
    plot.title    = element_text(color = text_col, face = "bold",
                                 size = 12, margin = margin(b = 4)),
    plot.subtitle = element_text(color = subtitle_col, size = 9,
                                 margin = margin(b = 8)),
    legend.position    = "bottom",
    legend.direction   = "horizontal",
    legend.title       = element_text(color = text_col, size = 9, face = "bold"),
    legend.text        = element_text(color = text_col, size = 8),
    legend.background  = element_rect(fill = bg_col, color = NA),
    legend.key         = element_rect(fill = bg_col, color = NA),
    legend.key.width   = unit(1.5, "cm"),
    legend.key.height  = unit(0.35, "cm"),
    plot.margin = margin(6, 8, 6, 8)
  )

# Download CA state boundary from Census Bureau; fall back to convex hull of
# conserved areas if tigris is unavailable or the download fails.
ca_outline_layer <- tryCatch({
  ca_state <- tigris::states(cb = TRUE, progress_bar = FALSE) |>
    filter(STUSPS == "CA") |>
    st_transform(crs = 3310)
  list(
    geom_sf(data = ca_state, fill = NA, color = border_col, linewidth = 0.4),
    coord_sf(expand = FALSE)
  )
}, error = function(e) {
  ca_hull <- st_union(conserved |> st_make_valid()) |> st_convex_hull()
  list(
    geom_sf(data = ca_hull, fill = NA, color = border_col, linewidth = 0.4),
    coord_sf(expand = FALSE)
  )
})

# Light warm grey: neutral "land cover" feel; does not compete with data scales.
conserved_base <- geom_sf(
  data  = conserved_gap12,
  fill  = "#c8cfc8",
  color = NA,
  alpha = 0.80
)
```

### Map helper function

Rather than repeating nearly identical `ggplot()` calls for each gap type and each sensitivity threshold, a single helper function takes the data and aesthetic column names as arguments. This keeps the map-building loop concise and ensures all panels are formatted identically.

```{r}
#| label: map-helper

# Builds a two-layer gap map (non-gap cells beneath gap cells).
# `data`         — sf object with gap and priority columns
# `gap_col`      — character: name of the logical gap column (e.g. "rich_gap_t")
# `priority_col` — character: name of the four-category label column
# `colors`       — named character vector mapping priority labels to hex colours
# `title`, `subtitle` — map annotation strings
make_gap_map <- function(data, gap_col, priority_col, colors, title, subtitle) {
  ggplot() +
    conserved_base +
    ca_outline_layer +
    # Non-gap cells rendered first so gap cells always sit on top.
    geom_sf(
      data  = data |> filter(!.data[[gap_col]]),
      aes(color = .data[[priority_col]], fill = .data[[priority_col]]),
      size = 0.1, alpha = 0.65, linewidth = 0
    ) +
    geom_sf(
      data  = data |> filter(.data[[gap_col]]),
      aes(color = .data[[priority_col]], fill = .data[[priority_col]]),
      size = 0.1, alpha = 1.0, linewidth = 0
    ) +
    scale_color_manual(
      values = colors, name = NULL,
      guide  = guide_legend(override.aes = list(size = 3),
                            nrow = 2, title.position = "top")
    ) +
    scale_fill_manual(
      values = colors, name = NULL,
      guide  = guide_legend(override.aes = list(size = 3),
                            nrow = 2, title.position = "top")
    ) +
    labs(title = title, subtitle = subtitle) +
    map_theme
}
```

------------------------------------------------------------------------

## Maps

### Fig. 1 — Conservation Gap: High Richness, Unprotected

Cells are split into four categories: high/lower richness crossed with protected/unprotected. The key category — high richness AND unprotected (orange) — represents locations where the conservation network fails to capture above-average bee diversity.

```{r}
#| label: fig-richness-gap
#| fig-alt: "Conservation gap map showing high bee richness areas outside GAP 1 & 2 lands"

gap_colors_rich <- c(
  "High richness, unprotected"  = "#D55E00",   # vermillion (Wong 2011)
  "High richness, protected"    = "#0072B2",   # blue
  "Lower richness, unprotected" = "#BBBBBB",
  "Lower richness, protected"   = "#888888"
)

fig1 <- make_gap_map(
  data         = bee_gaps,
  gap_col      = "rich_gap",
  priority_col = "rich_priority",
  colors       = gap_colors_rich,
  title        = "Fig. 1 — Conservation Gap: Bee Species Richness",
  subtitle     = glue(
    "Orange = high richness (≥{round(rich_thresh,1)} spp) AND outside protected lands  ·  ",
    "Threshold = {gap_quantile*100}th percentile"
  )
)

fig1
```

### Fig. 2 — Conservation Gap: High Endemism, Unprotected

CWE (corrected weighted endemism) upweights species with restricted geographic ranges. Areas with moderate total richness can still harbour many narrow-range endemics, so the spatial pattern of endemism gaps may differ substantially from richness gaps. The same colour scheme as Fig. 1 allows direct comparison.

```{r}
#| label: fig-cwe-gap
#| fig-alt: "Conservation gap map showing high bee endemism areas outside GAP 1 & 2 lands"

gap_colors_cwe <- c(
  "High endemism, unprotected"  = "#D55E00",
  "High endemism, protected"    = "#0072B2",
  "Lower endemism, unprotected" = "#BBBBBB",
  "Lower endemism, protected"   = "#888888"
)

fig2 <- make_gap_map(
  data         = bee_gaps,
  gap_col      = "cwe_gap",
  priority_col = "cwe_priority",
  colors       = gap_colors_cwe,
  title        = "Fig. 2 — Conservation Gap: Bee Rarity (CWE)",
  subtitle     = glue(
    "Orange = high rarity (≥{round(cwe_thresh,4)} CWE) AND outside protected lands  ·  ",
    "Threshold = {gap_quantile*100}th percentile"
  )
)

fig2
```

### Fig. 3 — Dual-Gap Priority Cells

Cells that are simultaneously above the richness threshold AND the endemism threshold AND outside GAP 1 & 2 are the highest-priority conservation gaps — protecting them would advance two independent objectives at once. A five-category palette distinguishes dual-gap cells (red) from single-dimension gaps and already-protected high-diversity cells.

```{r}
#| label: fig-dual-gap
#| fig-alt: "Map showing cells high in both bee richness and endemism but outside GAP 1 & 2 lands"

bee_gaps <- bee_gaps |>
  mutate(
    dual_gap = rich_gap & cwe_gap,
    dual_priority = case_when(
      dual_gap      ~ "High richness + endemism, unprotected",
      rich_gap      ~ "High richness only, unprotected",
      cwe_gap       ~ "High endemism only, unprotected",
      is_protected  ~ "Protected (GAP 1 & 2)",
      TRUE          ~ "Lower diversity, unprotected"
    )
  )

n_dual <- sum(bee_gaps$dual_gap, na.rm = TRUE)
message(glue(
  "{n_dual} cells are high in BOTH richness AND endemism and are unprotected."
))

dual_colors <- c(
  "High richness + endemism, unprotected" = "#CC0000",
  "High richness only, unprotected"       = "#E69F00",
  "High endemism only, unprotected"       = "#56B4E9",
  "Protected (GAP 1 & 2)"                = "#009E73",
  "Lower diversity, unprotected"          = "#DDDDDD"
)

# The dual-gap map has three priority tiers, so it is built directly rather
# than through make_gap_map() which assumes only two rendering layers.
fig3 <- ggplot() +
  conserved_base +
  ca_outline_layer +
  geom_sf(
    data  = bee_gaps |> filter(!rich_gap & !cwe_gap),
    aes(color = dual_priority, fill = dual_priority),
    size = 0.8, alpha = 0.5, linewidth = 0
  ) +
  geom_sf(
    data  = bee_gaps |> filter((rich_gap | cwe_gap) & !dual_gap),
    aes(color = dual_priority, fill = dual_priority),
    size = 1.2, alpha = 0.8, linewidth = 0
  ) +
  geom_sf(
    data  = bee_gaps |> filter(dual_gap),
    aes(color = dual_priority, fill = dual_priority),
    size = 2.0, alpha = 1.0, linewidth = 0
  ) +
  scale_color_manual(
    values = dual_colors, name = NULL,
    guide  = guide_legend(override.aes = list(size = 3),
                          nrow = 3, title.position = "top")
  ) +
  scale_fill_manual(
    values = dual_colors, name = NULL,
    guide  = guide_legend(override.aes = list(size = 3),
                          nrow = 3, title.position = "top")
  ) +
  labs(
    title    = "Fig. 3 — Dual Conservation Gap: High Bee Richness AND Endemism, Unprotected",
    subtitle = glue(
      "Red = {n_dual} cells unprotected AND high in both richness ",
      "(≥{round(rich_thresh,1)} spp) AND endemism (CWE ≥{round(cwe_thresh,4)})  ·  ",
      "Threshold = {gap_quantile*100}th percentile"
    )
  ) +
  map_theme

fig3
```

------------------------------------------------------------------------

## Summary Statistics

The three tables below quantify the gap patterns shown in Figs. 1–3. Any change to `gap_quantile` in the gap-thresholds chunk automatically propagates through both the maps and these tables.

### Table 1 — Richness gap

```{r}
#| label: table1-richness-summary

bee_gaps |>
  st_drop_geometry() |>
  group_by(rich_priority) |>
  summarise(
    n_cells         = n(),
    mean_richness   = round(mean(richness,   na.rm = TRUE), 1),
    median_richness = round(median(richness, na.rm = TRUE), 1)
  ) |>
  arrange(desc(mean_richness)) |>
  knitr::kable(caption = "Table 1. Bee richness by conservation status")
```

### Table 2 — Endemism gap

```{r}
#| label: table2-cwe-summary

bee_gaps |>
  st_drop_geometry() |>
  group_by(cwe_priority) |>
  summarise(
    n_cells    = n(),
    mean_cwe   = round(mean(cwe,   na.rm = TRUE), 4),
    median_cwe = round(median(cwe, na.rm = TRUE), 4)
  ) |>
  arrange(desc(mean_cwe)) |>
  knitr::kable(caption = "Table 2. Bee endemism (CWE) by conservation status")
```

### Table 3 — Dual-gap overlap

```{r}
#| label: table3-dual-summary

dual_gap <- bee_gaps |>
  st_drop_geometry() |>
  filter(rich_gap & cwe_gap)

cat(glue(
  "{nrow(dual_gap)} grid cells are high in BOTH richness AND endemism ",
  "and fall outside GAP 1 & 2 protected areas — these are the ",
  "highest-priority gaps.\n"
))

dual_gap |>
  summarise(
    n_dual_gap_cells = n(),
    mean_richness    = round(mean(richness, na.rm = TRUE), 1),
    mean_cwe         = round(mean(cwe,      na.rm = TRUE), 4)
  ) |>
  knitr::kable(
    caption   = "Table 3. Dual-gap cells: high in both richness and endemism, unprotected",
    col.names = c("Dual-gap cells (n)", "Mean richness", "Mean CWE")
  )
```

------------------------------------------------------------------------

## Sensitivity Analysis: Areal Overlap Thresholds

The main analysis classifies a cell as "protected" if it intersects GAP 1 & 2 land at all (any overlap). In practice a cell might overlap only at a corner. This section tests gap counts under stricter minimum areal overlap thresholds (10%, 25%, 50%, 75%).

### Compute overlap areas

The intersection is run once here; cells are reclassified under each threshold in the next chunk.

```{r}
#| label: build-bee-polygons
#| cache: true

# bee_prot already carries square polygon geometry from grid-to-squares.
# We confirm this here and compute each cell's area for the proportion calc.
bee_cells <- bee_prot
bee_cells$cell_area <- as.numeric(st_area(bee_cells))

message(glue(
  "Cell area range: {round(min(bee_cells$cell_area)/1e6, 1)} – ",
  "{round(max(bee_cells$cell_area)/1e6, 1)} km²"
))
```

```{r}
#| label: compute-overlap-area
#| cache: true

# st_intersection clips each bee cell to the GAP 1/2 union; the resulting
# geometry's area is the actual overlap between that cell and protected land.
bee_gap12_intersections <- suppressWarnings(suppressMessages(
  st_intersection(
    bee_cells[, c("bee_cell_id", "cell_area")],
    conserved_union["protected"]
  )
))

if (nrow(bee_gap12_intersections) > 0) {
  bee_gap12_intersections$overlap_area <- as.numeric(
    st_area(bee_gap12_intersections)
  )
  overlap_by_cell <- bee_gap12_intersections |>
    st_drop_geometry() |>
    group_by(bee_cell_id) |>
    summarise(overlap_area = sum(overlap_area, na.rm = TRUE), .groups = "drop")
} else {
  overlap_by_cell <- tibble(bee_cell_id = integer(), overlap_area = numeric())
}

# Join overlap back to bee cells and compute the proportion (capped at 1.0).
bee_overlap <- bee_cells |>
  left_join(overlap_by_cell, by = "bee_cell_id") |>
  mutate(
    overlap_area = replace_na(overlap_area, 0),
    overlap_prop = pmin(overlap_area / cell_area, 1)
  )

message(glue(
  "Overlap proportion — min: {round(min(bee_overlap$overlap_prop), 3)}, ",
  "median: {round(median(bee_overlap$overlap_prop), 3)}, ",
  "max: {round(max(bee_overlap$overlap_prop), 3)}"
))
```

### Build sensitivity summary tables

```{r}
#| label: sensitivity-summary

overlap_thresholds <- c(0.10, 0.25, 0.50, 0.75)

# Reclassify every cell under each threshold and stack into one long data frame.
bee_overlap_long <- map_dfr(overlap_thresholds, function(thresh) {
  bee_overlap |>
    st_drop_geometry() |>
    mutate(
      overlap_threshold   = thresh,
      threshold_label     = scales::percent(thresh, accuracy = 1),
      is_protected_thresh = overlap_prop >= thresh,
      rich_gap_thresh     = richness >= rich_thresh & !is_protected_thresh,
      cwe_gap_thresh      = cwe      >= cwe_thresh  & !is_protected_thresh,
      dual_gap_thresh     = rich_gap_thresh & cwe_gap_thresh
    )
})

# Table S1a: protection status by threshold
prot_summary_sensitivity <- bee_overlap_long |>
  group_by(overlap_threshold, threshold_label) |>
  summarise(
    n_cells_total       = n(),
    n_cells_protected   = sum(is_protected_thresh, na.rm = TRUE),
    n_cells_unprotected = sum(!is_protected_thresh, na.rm = TRUE),
    pct_cells_protected = round(100 * mean(is_protected_thresh, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  arrange(overlap_threshold)

# Table S1b: gap counts by threshold
gap_summary_sensitivity <- bee_overlap_long |>
  group_by(overlap_threshold, threshold_label) |>
  summarise(
    rich_gap_cells = sum(rich_gap_thresh,  na.rm = TRUE),
    cwe_gap_cells  = sum(cwe_gap_thresh,   na.rm = TRUE),
    dual_gap_cells = sum(dual_gap_thresh,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(overlap_threshold)

# Merge into a single wide table for display and export.
combined_sensitivity <- prot_summary_sensitivity |>
  left_join(gap_summary_sensitivity,
            by = c("overlap_threshold", "threshold_label")) |>
  transmute(
    `Overlap threshold`  = threshold_label,
    `Total cells`        = n_cells_total,
    `Protected cells`    = n_cells_protected,
    `Unprotected cells`  = n_cells_unprotected,
    `% Protected`        = pct_cells_protected,
    `Richness gap cells` = rich_gap_cells,
    `Endemism gap cells` = cwe_gap_cells,
    `Dual-gap cells`     = dual_gap_cells
  )
```

### Table S1 — Sensitivity of gap counts to overlap threshold

```{r}
#| label: tableS1-sensitivity-display

knitr::kable(
  combined_sensitivity,
  caption = paste0(
    "Table S1. Sensitivity of conservation gap counts to minimum areal overlap ",
    "threshold with GAP Status 1 & 2 protected lands. ",
    "A cell is classified as 'protected' only if the stated proportion of its ",
    "15 × 15 km area overlaps with GAP 1 or 2 lands. ",
    "'Richness gap' = high species richness (≥ 75th percentile) and unprotected. ",
    "'Endemism gap' = high corrected weighted endemism (CWE; ≥ 75th percentile) ",
    "and unprotected. ",
    "'Dual-gap' = high in both richness and endemism and unprotected."
  ),
  align       = c("l", rep("r", 7)),
  format.args = list(big.mark = ",")
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width        = TRUE,
    font_size         = 12
  ) |>
  add_header_above(
    c(" " = 1, "Cell counts" = 4, "Conservation gaps" = 3),
    bold = TRUE
  ) |>
  row_spec(0, bold = TRUE) |>
  kableExtra::footnote(
    general = paste0(
      "Bee grid cells are 15 × 15 km. ",
      "GAP 1 & 2 protected lands from CA Nature Terrestrial 30×30 Conserved Areas ",
      "(GreenInfo Network / USGS PAD-US, 2024). ",
      "Biodiversity thresholds: species richness ≥ ",
      round(rich_thresh, 1), " spp and CWE ≥ ",
      round(cwe_thresh, 4),
      " (75th percentile across all grid cells)."
    ),
    general_title     = "Note:",
    footnote_as_chunk = TRUE
  )
```

### Export Table S1

```{r}
#| label: export-tableS1

fig_dir <- file.path("..", "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(
  combined_sensitivity,
  file.path(fig_dir, "table_S1_sensitivity.csv"),
  row.names = FALSE
)

# The CSV can be opened in Word or Excel and formatted with a journal table
# style. The styled HTML version above serves as the analytical record.
message("Table S1 exported to figures/table_S1_sensitivity.csv")
```

------------------------------------------------------------------------

## Protection Coverage of High-Biodiversity Cells

This section asks the inverse of the gap question: of all cells that meet the high-richness, high-endemism, or dual threshold, what fraction already falls inside GAP 1 & 2 lands? Results are shown for the binary intersection definition (main analysis) and for each areal-overlap threshold.

### Table 4 — Protection coverage

```{r}
#| label: table4-protection-coverage

# Main analysis: any intersection = protected
coverage_main <- bee_gaps |>
  st_drop_geometry() |>
  summarise(
    n_high_rich        = sum(richness >= rich_thresh, na.rm = TRUE),
    n_high_rich_prot   = sum(richness >= rich_thresh &  is_protected, na.rm = TRUE),
    n_high_rich_unprot = sum(richness >= rich_thresh & !is_protected, na.rm = TRUE),
    pct_high_rich_prot = round(100 * n_high_rich_prot / n_high_rich, 1),

    n_high_cwe        = sum(cwe >= cwe_thresh, na.rm = TRUE),
    n_high_cwe_prot   = sum(cwe >= cwe_thresh &  is_protected, na.rm = TRUE),
    n_high_cwe_unprot = sum(cwe >= cwe_thresh & !is_protected, na.rm = TRUE),
    pct_high_cwe_prot = round(100 * n_high_cwe_prot / n_high_cwe, 1),

    n_dual        = sum(richness >= rich_thresh & cwe >= cwe_thresh, na.rm = TRUE),
    n_dual_prot   = sum(richness >= rich_thresh & cwe >= cwe_thresh &  is_protected, na.rm = TRUE),
    n_dual_unprot = sum(richness >= rich_thresh & cwe >= cwe_thresh & !is_protected, na.rm = TRUE),
    pct_dual_prot = round(100 * n_dual_prot / n_dual, 1)
  ) |>
  mutate(threshold = "Any overlap (main analysis)", .before = 1)

# Sensitivity thresholds
coverage_sensitivity <- map_dfr(c(0.10, 0.25, 0.50, 0.75), function(thresh) {
  bee_overlap |>
    st_drop_geometry() |>
    mutate(is_protected_t = overlap_prop >= thresh) |>
    summarise(
      n_high_rich        = sum(richness >= rich_thresh, na.rm = TRUE),
      n_high_rich_prot   = sum(richness >= rich_thresh &  is_protected_t, na.rm = TRUE),
      n_high_rich_unprot = sum(richness >= rich_thresh & !is_protected_t, na.rm = TRUE),
      pct_high_rich_prot = round(100 * n_high_rich_prot / n_high_rich, 1),

      n_high_cwe        = sum(cwe >= cwe_thresh, na.rm = TRUE),
      n_high_cwe_prot   = sum(cwe >= cwe_thresh &  is_protected_t, na.rm = TRUE),
      n_high_cwe_unprot = sum(cwe >= cwe_thresh & !is_protected_t, na.rm = TRUE),
      pct_high_cwe_prot = round(100 * n_high_cwe_prot / n_high_cwe, 1),

      n_dual        = sum(richness >= rich_thresh & cwe >= cwe_thresh, na.rm = TRUE),
      n_dual_prot   = sum(richness >= rich_thresh & cwe >= cwe_thresh &  is_protected_t, na.rm = TRUE),
      n_dual_unprot = sum(richness >= rich_thresh & cwe >= cwe_thresh & !is_protected_t, na.rm = TRUE),
      pct_dual_prot = round(100 * n_dual_prot / n_dual, 1)
    ) |>
    mutate(threshold = glue("≥{scales::percent(thresh, accuracy = 1)} overlap"),
           .before = 1)
})

coverage_table <- bind_rows(coverage_main, coverage_sensitivity) |>
  transmute(
    `Protection threshold`    = threshold,
    `High richness cells (n)` = n_high_rich,
    `Protected (n)`           = n_high_rich_prot,
    `Unprotected (n)`         = n_high_rich_unprot,
    `% Protected`             = pct_high_rich_prot,
    `High endemism cells (n)` = n_high_cwe,
    `Protected (n) `          = n_high_cwe_prot,    # trailing space avoids duplicate col name
    `Unprotected (n) `        = n_high_cwe_unprot,
    `% Protected `            = pct_high_cwe_prot,
    `Dual high cells (n)`     = n_dual,
    `Protected (n)  `         = n_dual_prot,
    `Unprotected (n)  `       = n_dual_unprot,
    `% Protected  `           = pct_dual_prot
  )

knitr::kable(
  coverage_table,
  caption = paste0(
    "Table 4. Percentage of high-biodiversity bee grid cells covered by ",
    "GAP Status 1 & 2 protected lands under California's 30×30 initiative, ",
    "by protection definition. ",
    "'High richness' = ≥ ", round(rich_thresh, 1), " species (75th percentile). ",
    "'High endemism' = CWE ≥ ", round(cwe_thresh, 4), " (75th percentile). ",
    "'Dual high' = above both thresholds simultaneously."
  ),
  align = c("l", rep("r", 12)),
  format.args = list(big.mark = ",")
) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width        = TRUE,
    font_size         = 12
  ) |>
  add_header_above(
    c(" " = 1,
      "High richness" = 4,
      "High endemism" = 4,
      "Dual high (richness + endemism)" = 4),
    bold = TRUE
  ) |>
  row_spec(0, bold = TRUE) |>
  row_spec(1, background = "#eaf2fb") |>
  kableExtra::footnote(
    general = paste0(
      "Grid cells are 15 × 15 km. ",
      "GAP 1 & 2 lands from CA Nature Terrestrial 30×30 Conserved Areas ",
      "(GreenInfo Network / USGS PAD-US, 2024). ",
      "The main analysis row ('Any overlap') classifies a cell as protected ",
      "if it intersects GAP 1 & 2 land at all. ",
      "Subsequent rows require the stated minimum proportion of the cell area ",
      "to overlap with protected land before the cell is counted as protected."
    ),
    general_title     = "Note:",
    footnote_as_chunk = TRUE
  )

write.csv(
  coverage_table,
  file.path(fig_dir, "table_4_protection_coverage.csv"),
  row.names = FALSE
)
message("Table 4 exported to figures/table_4_protection_coverage.csv")
```

------------------------------------------------------------------------

## Sensitivity Maps

Gap maps (richness, endemism, dual) are produced for each of the four areal-overlap thresholds. All formatting follows Figs. 1–3 exactly; the only change across panels is which cells are classified as "protected."

```{r}
#| label: figs-sensitivity-thresholds
#| fig-width: 12
#| fig-height: 8
#| warning: false
#| message: false

thresh_labels <- tibble(
  overlap_threshold = c(0.10, 0.25, 0.50, 0.75),
  threshold_label   = c("10%", "25%", "50%", "75%")
)

# Storage lists for optional downstream export
rich_maps <- list()
cwe_maps  <- list()
dual_maps <- list()

for (i in seq_len(nrow(thresh_labels))) {

  thresh       <- thresh_labels$overlap_threshold[i]
  thresh_label <- thresh_labels$threshold_label[i]

  # Reclassify protection and gap status at this threshold.
  # Geometry is dropped here and re-attached from bee.data (same row order).
  thresh_data <- bee_overlap |>
    st_drop_geometry() |>
    mutate(
      is_protected_t = overlap_prop >= thresh,

      rich_gap_t  = richness >= rich_thresh & !is_protected_t,
      cwe_gap_t   = cwe      >= cwe_thresh  & !is_protected_t,
      dual_gap_t  = rich_gap_t & cwe_gap_t,

      rich_priority_t = case_when(
        richness >= rich_thresh & !is_protected_t ~ "High richness, unprotected",
        richness >= rich_thresh &  is_protected_t ~ "High richness, protected",
        richness <  rich_thresh & !is_protected_t ~ "Lower richness, unprotected",
        TRUE                                      ~ "Lower richness, protected"
      ),
      cwe_priority_t = case_when(
        cwe >= cwe_thresh & !is_protected_t ~ "High endemism, unprotected",
        cwe >= cwe_thresh &  is_protected_t ~ "High endemism, protected",
        cwe <  cwe_thresh & !is_protected_t ~ "Lower endemism, unprotected",
        TRUE                                ~ "Lower endemism, protected"
      ),
      dual_priority_t = case_when(
        dual_gap_t     ~ "High richness + endemism, unprotected",
        rich_gap_t     ~ "High richness only, unprotected",
        cwe_gap_t      ~ "High endemism only, unprotected",
        is_protected_t ~ "Protected (GAP 1 & 2)",
        TRUE           ~ "Lower diversity, unprotected"
      )
    )

  thresh_sf <- bee.data |>
    select(geom) |>
    bind_cols(thresh_data)

  n_rich_gap <- sum(thresh_sf$rich_gap_t, na.rm = TRUE)
  n_cwe_gap  <- sum(thresh_sf$cwe_gap_t,  na.rm = TRUE)
  n_dual_gap <- sum(thresh_sf$dual_gap_t, na.rm = TRUE)

  message(glue(
    "Threshold {thresh_label}: richness gap = {n_rich_gap}, ",
    "CWE gap = {n_cwe_gap}, dual gap = {n_dual_gap}"
  ))

  # Figs. S1a / S2a / S3a / S4a — richness gap at this threshold
  map_rich <- make_gap_map(
    data         = thresh_sf,
    gap_col      = "rich_gap_t",
    priority_col = "rich_priority_t",
    colors       = gap_colors_rich,
    title        = glue("Richness Gap — {thresh_label} overlap threshold"),
    subtitle     = glue(
      "Orange = high richness (≥{round(rich_thresh,1)} spp) AND ",
      "< {thresh_label} overlap with GAP 1 & 2  ·  n gap cells = {n_rich_gap}"
    )
  )

  # Figs. S1b / S2b / S3b / S4b — endemism gap at this threshold
  map_cwe <- make_gap_map(
    data         = thresh_sf,
    gap_col      = "cwe_gap_t",
    priority_col = "cwe_priority_t",
    colors       = gap_colors_cwe,
    title        = glue("Endemism Gap — {thresh_label} overlap threshold"),
    subtitle     = glue(
      "Orange = high CWE (≥{round(cwe_thresh,4)}) AND ",
      "< {thresh_label} overlap with GAP 1 & 2  ·  n gap cells = {n_cwe_gap}"
    )
  )

  # Figs. S1c / S2c / S3c / S4c — dual gap at this threshold (three layers)
  map_dual <- ggplot() +
    conserved_base +
    ca_outline_layer +
    geom_sf(
      data  = thresh_sf |> filter(!rich_gap_t & !cwe_gap_t),
      aes(color = dual_priority_t, fill = dual_priority_t),
      size = 0.1, alpha = 0.5, linewidth = 0
    ) +
    geom_sf(
      data  = thresh_sf |> filter((rich_gap_t | cwe_gap_t) & !dual_gap_t),
      aes(color = dual_priority_t, fill = dual_priority_t),
      size = 0.1, alpha = 0.8, linewidth = 0
    ) +
    geom_sf(
      data  = thresh_sf |> filter(dual_gap_t),
      aes(color = dual_priority_t, fill = dual_priority_t),
      size = 0.1, alpha = 1.0, linewidth = 0
    ) +
    scale_color_manual(
      values = dual_colors, name = NULL,
      guide  = guide_legend(override.aes = list(size = 3),
                            nrow = 3, title.position = "top")
    ) +
    scale_fill_manual(
      values = dual_colors, name = NULL,
      guide  = guide_legend(override.aes = list(size = 3),
                            nrow = 3, title.position = "top")
    ) +
    labs(
      title    = glue("Dual Gap — {thresh_label} overlap threshold"),
      subtitle = glue(
        "Red = {n_dual_gap} cells high in richness AND endemism, ",
        "< {thresh_label} overlap with GAP 1 & 2"
      )
    ) +
    map_theme

  print(map_rich)
  print(map_cwe)
  print(map_dual)

  rich_maps[[thresh_label]] <- map_rich
  cwe_maps[[thresh_label]]  <- map_cwe
  dual_maps[[thresh_label]] <- map_dual
}
```

------------------------------------------------------------------------

## Export

### Export manuscript figures (Figs. 1–3)

```{r}
#| label: export-main-figures

fig_specs <- list(
  list(plot = fig1, name = "fig1_richness_gap"),
  list(plot = fig2, name = "fig2_cwe_gap"),
  list(plot = fig3, name = "fig3_dual_gap")
)

for (fig in fig_specs) {
  ggsave(
    filename    = file.path(fig_dir, paste0(fig$name, ".tiff")),
    plot        = fig$plot,
    width       = 174,
    height      = 160,
    units       = "mm",
    dpi         = 600,
    compression = "lzw"
  )
  message(glue("Saved: {fig$name}.tiff"))
}

message(glue("Main figures exported to: {normalizePath(fig_dir)}"))
```

### Export sensitivity maps (Figs. S1–S4)

```{r}
#| label: export-sensitivity-maps

map_lists   <- list(richness = rich_maps, endemism = cwe_maps, dual = dual_maps)
fig_numbers <- list(richness = "figS_richness", endemism = "figS_endemism",
                    dual = "figS_dual")

for (metric in names(map_lists)) {
  for (thresh_label in names(map_lists[[metric]])) {
    thresh_stem <- gsub("%", "pct", thresh_label)
    fname <- glue("{fig_numbers[[metric]]}_sens{thresh_stem}.tiff")
    ggsave(
      filename    = file.path(fig_dir, fname),
      plot        = map_lists[[metric]][[thresh_label]],
      width       = 174,
      height      = 160,
      units       = "mm",
      dpi         = 600,
      compression = "lzw"
    )
    message(glue("Saved: {fname}"))
  }
}

message(glue("All sensitivity maps exported to: {normalizePath(fig_dir)}"))
```

### Export gap layers (GeoPackage)

```{r}
#| label: export-gpkg

out_dir <- "outputs"
dir.create(out_dir, showWarnings = FALSE)

# Full classified grid — all cells with gap-status columns
st_write(bee_gaps, file.path(out_dir, "bee_conservation_gaps.gpkg"),
         append = FALSE)

# Priority subset — only cells that are a gap on at least one dimension
priority_gaps <- bee_gaps |> filter(rich_gap | cwe_gap)

st_write(priority_gaps, file.path(out_dir, "bee_priority_gaps.gpkg"),
         append = FALSE)

message(glue(
  "Exported {nrow(bee_gaps)} classified cells and ",
  "{nrow(priority_gaps)} priority gap cells to {out_dir}/"
))
```

------------------------------------------------------------------------

## Data Integrity Validation

Each check is assigned one of three statuses:

-   **FAIL** — incorrect results; must be resolved before reporting findings.
-   **WARN** — may be acceptable; document reasoning if proceeding.
-   **PASS** — no problem detected.

The helper functions (`pass`, `warn`, `fail`) were defined in the `setup` chunk so they are available throughout the document.

```{r}
#| label: validation
#| message: true

# Reset the checks accumulator before running (guards against re-runs).
checks <- tibble(
  severity    = character(),
  id          = character(),
  description = character(),
  detail      = character()
)

# ── Check 1: No NA coordinates in bee.data after loading ──────────────────────
{
  n_na_coords <- if (inherits(st_geometry(bee.data), "sfc_POINT")) {
    coords <- st_coordinates(bee.data)
    sum(is.na(coords[, 1]) | is.na(coords[, 2]))
  } else {
    0L  # polygon geometry: coordinate NAs not directly checkable this way
  }
  if (n_na_coords > 0) {
    fail("1-coord-na", "NA coordinates in bee.data",
         glue("{n_na_coords} cells with NA x or y"))
  } else {
    pass("1-coord-na", "No NA coordinates in bee.data",
         glue("{nrow(bee.data)} cells loaded"))
  }
}

# ── Check 2: No NA values in richness or cwe ──────────────────────────────────
# NA richness/cwe causes case_when() to silently assign cells to the catch-all
# category regardless of actual protection status.
{
  n_na_rich    <- sum(is.na(bee_prot$richness))
  n_na_cwe     <- sum(is.na(bee_prot$cwe))
  n_na_richgap <- sum(is.na(bee_gaps$rich_gap))
  n_na_cwegap  <- sum(is.na(bee_gaps$cwe_gap))

  if (n_na_rich > 0)    fail("2-na-richness",  "NA in richness",        glue("{n_na_rich} cells"))   else pass("2-na-richness",  "No NA in richness")
  if (n_na_cwe > 0)     fail("2-na-cwe",        "NA in cwe",             glue("{n_na_cwe} cells"))    else pass("2-na-cwe",        "No NA in cwe")
  if (n_na_richgap > 0 || n_na_cwegap > 0) {
    fail("2-na-gap-flags", "NA in rich_gap or cwe_gap",
         glue("rich_gap: {n_na_richgap}, cwe_gap: {n_na_cwegap}"))
  } else {
    pass("2-na-gap-flags", "No NA in rich_gap or cwe_gap")
  }
}

# ── Check 3: Row count unchanged after spatial join ───────────────────────────
{
  if (nrow(bee_prot) != nrow(bee.data)) {
    fail("3-join-dedup", "Row count changed after spatial join",
         glue("bee.data: {nrow(bee.data)}, bee_prot: {nrow(bee_prot)}"))
  } else {
    pass("3-join-dedup", "Row count unchanged after spatial join",
         glue("{nrow(bee_prot)} rows"))
  }
}

# ── Check 4: Gap1_acres and Gap2_acres are numeric ────────────────────────────
# If stored as character, filter(> 0) evaluates to NA and conserved_gap12
# may be empty without any error.
{
  cls1 <- class(conserved$Gap1_acres)
  cls2 <- class(conserved$Gap2_acres)
  ok   <- c("numeric", "integer", "double")
  if (!cls1 %in% ok) fail("4-acres-type", "Gap1_acres not numeric", glue("class: {cls1}")) else pass("4-acres-type", "Gap1/2_acres are numeric", glue("{cls1}, {cls2}"))
  if (nrow(conserved_gap12) == 0) fail("4-gap12-empty", "conserved_gap12 is empty") else pass("4-gap12-empty", "conserved_gap12 non-empty", glue("{nrow(conserved_gap12)} polygons"))
}

# ── Check 5: No zero-area or NaN cells ────────────────────────────────────────
# Zero cell_area causes division-by-zero → NaN in overlap_prop, which
# sum(..., na.rm = TRUE) silently excludes, understating unprotected cells.
{
  if (exists("bee_cells") && "cell_area" %in% names(bee_cells)) {
    n_zero <- sum(bee_cells$cell_area == 0, na.rm = TRUE)
    n_na   <- sum(is.na(bee_cells$cell_area))
    if (n_zero > 0) fail("5-zero-area", "Degenerate cells (area = 0)",  glue("{n_zero} cells")) else pass("5-zero-area", "No zero-area cells")
    if (n_na   > 0) fail("5-na-area",   "NA cell_area values",           glue("{n_na} cells"))   else pass("5-na-area",   "No NA cell_area values")
    if (exists("bee_overlap")) {
      n_nan <- sum(is.nan(st_drop_geometry(bee_overlap)$overlap_prop))
      if (n_nan > 0) fail("5-nan-prop", "NaN in overlap_prop", glue("{n_nan} cells")) else pass("5-nan-prop", "No NaN in overlap_prop")
    }
  } else {
    warn("5-zero-area", "bee_cells not found — sensitivity chunk may not have run")
  }
}

# ── Check 6: CRS is EPSG 3310 for bee.data and conserved_union ───────────────
{
  epsg_bee   <- st_crs(bee.data)$epsg
  epsg_union <- st_crs(conserved_union)$epsg
  if (is.na(epsg_bee))     warn("6-crs-bee",   "bee.data CRS has no EPSG code") else if (epsg_bee != 3310L)   fail("6-crs-bee",   "bee.data not EPSG 3310",   glue("Detected: {epsg_bee}"))   else pass("6-crs-bee",   "bee.data CRS = EPSG 3310")
  if (!isTRUE(epsg_union == 3310L)) fail("6-crs-union", "conserved_union not EPSG 3310", glue("Detected: {epsg_union}")) else pass("6-crs-union", "conserved_union CRS = EPSG 3310")
}

# ── Check 7: conserved_union geometry is valid and non-empty ─────────────────
{
  if (any(st_is_empty(conserved_union$geometry))) fail("7-union-empty", "conserved_union is empty") else pass("7-union-empty", "conserved_union is non-empty")
  if (!all(st_is_valid(conserved_union$geometry))) fail("7-union-valid", "conserved_union is invalid") else pass("7-union-valid", "conserved_union is valid")
}

# ── Check 8: Row count unchanged after sensitivity left_join ─────────────────
{
  if (exists("bee_overlap") && exists("bee_cells")) {
    if (nrow(bee_overlap) != nrow(bee_cells)) {
      fail("8-overlap-rows", "Row count changed after sensitivity join",
           glue("bee_cells: {nrow(bee_cells)}, bee_overlap: {nrow(bee_overlap)}"))
    } else {
      pass("8-overlap-rows", "Row count unchanged after sensitivity join",
           glue("{nrow(bee_overlap)} rows"))
    }
  } else {
    warn("8-overlap-rows", "bee_overlap or bee_cells not found — sensitivity chunk may not have run")
  }
}

# ── Check 9: overlap_thresholds starts at 0.10 (not 0.00) ───────────────────
# The 0% row is intentionally absent from gap_summary_sensitivity.
# This WARN is expected and benign.
{
  if (exists("gap_summary_sensitivity")) {
    sens_dual_0 <- gap_summary_sensitivity |>
      filter(overlap_threshold == 0) |>
      pull(dual_gap_cells)
    if (length(sens_dual_0) == 0) {
      warn("9-sens-consistency",
           "No 0% threshold row in gap_summary_sensitivity — expected (thresholds start at 10%)",
           "Confirmed 8 Apr 2026")
    } else if (sens_dual_0 != dual_gap_n) {
      fail("9-sens-consistency", "0% dual-gap count ≠ main analysis dual_gap_n",
           glue("Main: {dual_gap_n}, Sensitivity 0%: {sens_dual_0}"))
    } else {
      pass("9-sens-consistency", "Sensitivity 0% dual-gap matches main analysis",
           glue("Both = {dual_gap_n}"))
    }
  } else {
    warn("9-sens-consistency", "gap_summary_sensitivity not found")
  }
}

# ── Check 10: YAML message suppression reminder ───────────────────────────────
{
  warn("10-message-suppressed",
       "YAML execute: message: false suppresses message() output in rendered HTML",
       "Use cat() for counts that must appear in the final document")
}

# ── Check 11: Single shapefile in zip ─────────────────────────────────────────
{
  if (exists("conserved_dir") && dir.exists(conserved_dir)) {
    all_shp <- list.files(conserved_dir, pattern = "\\.shp$",
                          full.names = TRUE, recursive = TRUE, ignore.case = TRUE)
    if (length(all_shp) > 1) {
      warn("11-multi-shp", "Multiple .shp files in zip — first was used",
           paste(basename(all_shp), collapse = ", "))
    } else if (length(all_shp) == 1) {
      pass("11-multi-shp", "Exactly one .shp in zip", basename(all_shp))
    } else {
      warn("11-multi-shp", "Temp dir cleaned — cannot recheck zip contents")
    }
  } else {
    warn("11-multi-shp", "conserved_dir no longer exists")
  }
}

# ── Check 12: No double-counting of acres ────────────────────────────────────
{
  n_both <- sum(conserved_gap12$Gap1_acres > 0 & conserved_gap12$Gap2_acres > 0,
                na.rm = TRUE)
  if (n_both > 0) {
    warn("12-acres-double", "Parcels with both Gap1 and Gap2 acreage > 0",
         glue("{n_both} parcels — acreage summary may overstate protected area"))
  } else {
    pass("12-acres-double", "No parcels with both Gap1 and Gap2 acreage > 0")
  }
}

# ── Check 13: dual_gap object exists and count matches gap-thresholds ─────────
{
  if (!exists("dual_gap")) {
    fail("13-dual-gap-exists", "Object 'dual_gap' not found — table3 chunk did not run")
  } else if (nrow(dual_gap) != dual_gap_n) {
    fail("13-dual-gap-count", "dual_gap count ≠ dual_gap_n",
         glue("dual_gap: {nrow(dual_gap)}, dual_gap_n: {dual_gap_n}"))
  } else {
    pass("13-dual-gap-count", "dual_gap count matches dual_gap_n",
         glue("n = {nrow(dual_gap)}"))
  }
}

# ── Print console summary ─────────────────────────────────────────────────────
cat("\n=======================================================================\n")
cat("  DATA INTEGRITY VALIDATION SUMMARY\n")
cat("=======================================================================\n\n")

n_fail <- sum(checks$severity == "FAIL")
n_warn <- sum(checks$severity == "WARN")
n_pass <- sum(checks$severity == "PASS")

ordered_checks <- bind_rows(
  filter(checks, severity == "FAIL"),
  filter(checks, severity == "WARN"),
  filter(checks, severity == "PASS")
)

for (i in seq_len(nrow(ordered_checks))) {
  row    <- ordered_checks[i, ]
  prefix <- switch(row$severity, "FAIL" = "[ FAIL ]", "WARN" = "[ WARN ]", "PASS" = "[ pass ]")
  cat(sprintf("%-10s  %s — %s\n", prefix, row$id, row$description))
  if (nchar(row$detail) > 0) {
    cat(paste(strwrap(row$detail, width = 68, prefix = "             "),
              collapse = "\n"), "\n")
  }
  cat("\n")
}

cat("-----------------------------------------------------------------------\n")
cat(sprintf("  %d FAIL   %d WARN   %d pass   (%d checks total)\n\n",
            n_fail, n_warn, n_pass, nrow(checks)))
cat(if (n_fail > 0) "  ACTION REQUIRED: resolve all FAIL items before reporting results.\n"
    else if (n_warn > 0) "  Review WARN items before submission.\n"
    else "  All checks passed. Proceed to submission.\n")
cat("=======================================================================\n\n")

# ── HTML table ────────────────────────────────────────────────────────────────
fail_rows <- which(ordered_checks$severity == "FAIL")
warn_rows <- which(ordered_checks$severity == "WARN")

# flextable masks kable_styling() and row_spec() — namespace explicitly.
ordered_checks |>
  select(severity, id, description, detail) |>
  knitr::kable(
    caption   = "Data integrity validation results",
    col.names = c("Status", "Check ID", "Description", "Detail"),
    align     = c("c", "l", "l", "l")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed", "bordered"),
    full_width        = TRUE,
    font_size         = 11
  ) |>
  row_spec(fail_rows, background = "#fde8e8", bold = TRUE) |>
  row_spec(warn_rows, background = "#fff8e1")
```

------------------------------------------------------------------------

## Notes on the 30×30 Conserved Areas Data

The Terrestrial 30×30 Conserved Areas layer (CA Nature, 2024) integrates:

-   **CPAD** — California Protected Areas Database (GreenInfo Network)
-   **CCED** — California Conservation Easement Database (GreenInfo Network)
-   **PAD-US** — USGS Gap Analysis Project (GAP status coding)

GAP status was assigned using an 80% areal overlap rule at the Superunit level. Areas with less than 80% overlap with any single GAP code were resolved at the Holdings level. Remaining uncoded areas were assigned GAP 4.

**Layer citation:** California Protected Areas Database (CPAD). December 2024. GreenInfo Network. www.calands.org
