---
title: "California Bee Diversity & Conservation Gaps"
author: "Gretchen LeBuhn"
date: "`r Sys.Date()`"
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 4
    theme: sandstone
    code-fold: true
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
---

## Overview

This document overlays California bee species richness and endemism (CWE) data
with the CA Nature Terrestrial 30×30 Conserved Areas layer to identify
**conservation gaps** — areas of high bee diversity or endemism that fall
outside GAP Status 1 & 2 protected lands.

Five maps are produced:

1. Bee richness overlaid on conserved lands
2. Bee endemism (CWE) overlaid on conserved lands
3. Conservation gaps — high richness, low protection
4. Conservation gaps — high endemism, low protection
5. Dual-gap priority cells — high in both richness and endemism, unprotected

A sensitivity analysis examines how gap counts change when the minimum
areal-overlap threshold required to classify a cell as "protected" is varied
from 0% to 50%.

------------------------------------------------------------------------

## Setup

All required packages are loaded here. `kableExtra` is included for
publication-quality table formatting in the sensitivity analysis.
The `data_dir` path points to the `bees_climate/data/` sibling directory
relative to this script.

```{r setup}
library(readr)       # fast CSV reading with clean column typing
library(dplyr)       # data manipulation verbs (filter, mutate, summarise)
library(tidyverse)   # meta-package: ggplot2, tidyr, stringr, purrr, etc.
library(sf)          # simple features: read, transform, and join spatial data
library(ggplot2)     # layer-by-layer map and figure construction
library(viridis)     # perceptually uniform, colorblind-safe colour palettes
library(patchwork)   # combine multiple ggplot objects into one figure
library(scales)      # format axis labels and legend text
library(glue)        # string interpolation for message() and labels
library(tigris)      # Census Bureau CA state boundary for map outlines
library(kableExtra)  # styled HTML tables with column spanners and footnotes

# All data files live in bees_climate/data/, one level above this script
data_dir <- file.path("..", "data")
```

------------------------------------------------------------------------

## Load Data

The two data-loading chunks below are independent of each other and are kept
separate so that each data source is clearly documented on its own. Both are
reprojected to CA Albers (EPSG 3310, units in metres) immediately on load so
that every subsequent spatial operation uses a consistent coordinate system.

### Bee diversity data

The processed bee data contains ecoregion assignments and grid-cell centroids
in CA Albers (EPSG 3310). The script prefers the GeoPackage (which preserves
geometry and CRS exactly) and falls back to the CSV if the GeoPackage is
absent.

```{r load-bee-data}
processed_dir <- file.path(data_dir, "processed")
gpkg_path     <- file.path(processed_dir, "eco.data.gpkg")
csv_path      <- file.path(processed_dir, "eco.data.csv")

if (file.exists(gpkg_path)) {

  bee.data <- st_read(gpkg_path, quiet = TRUE)
  message("Loaded bee data from GeoPackage: ", gpkg_path)

} else if (file.exists(csv_path)) {

  message("GeoPackage not found — loading from CSV: ", csv_path)

  bee.data <- read_csv(csv_path, show_col_types = FALSE) |>
    rename(x.axis = X_Axis_0_, y.axis = X_Axis_1_,
           cwe = X_ENDC_CWE_, richness = X_ENDC_RICHNESS_) |>
    select(x.axis, y.axis, cwe, richness,
           starts_with("bio"), starts_with("L")) |>
    drop_na(x.axis, y.axis) |>
    st_as_sf(coords = c("x.axis", "y.axis"), crs = 3310)

} else {
  stop(
    "Neither eco.data.gpkg nor eco.data.csv found in: ", processed_dir,
    "\nCheck that bees_climate/data/processed/ contains the bee data files."
  )
}

bee.data <- st_transform(bee.data, crs = 3310)

message(glue(
  "Bee data: {nrow(bee.data)} grid cells | ",
  "Richness range: {round(min(bee.data$richness, na.rm=TRUE), 1)} – ",
  "{round(max(bee.data$richness, na.rm=TRUE), 1)} | ",
  "CWE range: {round(min(bee.data$cwe, na.rm=TRUE), 3)} – ",
  "{round(max(bee.data$cwe, na.rm=TRUE), 3)}"
))
```

### 30×30 Conserved Areas shapefile

The CA Nature Terrestrial 30×30 layer is distributed as a zip archive.
The chunk below locates the zip in `bees_climate/data/`, extracts it to a
temporary directory, and reads the shapefile into R. Update
`conserved_zip_name` if your downloaded file has a different name.

```{r load-conserved}
conserved_zip_name <- "30x30_Conserved_Areas, Terrestrial_(2025).zip"
zip_path           <- file.path(data_dir, conserved_zip_name)

# Also try the URL-encoded variant (some systems save commas as %2C)
if (!file.exists(zip_path)) {
  zip_path_encoded <- file.path(
    data_dir,
    "30x30_Conserved_Areas%2C_Terrestrial_(2025).zip"
  )
  if (file.exists(zip_path_encoded)) {
    zip_path <- zip_path_encoded
  } else {
    stop(
      "Conserved areas zip not found.\n",
      "Expected at: ", file.path(data_dir, conserved_zip_name), "\n",
      "Check that bees_climate/data/ contains the 30x30 zip file."
    )
  }
}

message("Reading conserved areas from: ", zip_path)

conserved_dir <- tempfile("conserved_")
dir.create(conserved_dir)
unzip(zip_path, exdir = conserved_dir)

conserved_path <- list.files(
  conserved_dir,
  pattern     = "\\.shp$",
  full.names  = TRUE,
  recursive   = TRUE,
  ignore.case = TRUE
)

stopifnot(
  "No .shp file found after unzipping — check the zip contents" =
    length(conserved_path) >= 1
)

conserved_path <- conserved_path[1]
message("Reading shapefile from: ", conserved_path)

conserved <- st_read(conserved_path, quiet = TRUE) |>
  st_transform(crs = 3310)

message(glue("Conserved areas: {nrow(conserved)} polygons loaded."))
```

------------------------------------------------------------------------

## Prepare Conserved Areas

GAP Status codes indicate the management intent of a protected area:

- **GAP 1** — permanently protected, primary biodiversity focus
- **GAP 2** — managed mostly for biodiversity but may allow some extractive use
- **GAP 3** — multiple use, biodiversity not primary focus
- **GAP 4** — no known biodiversity focus (e.g. working lands)

This analysis defines "well conserved" as GAP 1 or GAP 2. Polygons with any
GAP 1 or GAP 2 acreage are filtered, their geometries are repaired with
`st_make_valid()`, and then dissolved into a single unified polygon with
`st_union()`. Dissolving before the spatial join (next chunk) is critical for
performance: joining against one polygon is orders of magnitude faster than
joining against tens of thousands of individual parcels.

```{r prep-conserved}
conserved_gap12 <- conserved |>
  filter(Gap1_acres > 0 | Gap2_acres > 0)

message(glue(
  "Invalid geometries before repair: {sum(!st_is_valid(conserved_gap12))}"
))

# Drop Z/M coordinates (some shapefiles carry elevation or measure values that
# can cause st_make_valid() to behave unexpectedly) then repair
conserved_gap12 <- conserved_gap12 |>
  mutate(geometry = st_zm(geometry, drop = TRUE, what = "ZM")) |>
  mutate(geometry = st_make_valid(geometry))

message(glue(
  "Invalid geometries after repair:  {sum(!st_is_valid(conserved_gap12))}"
))

message("Dissolving GAP 1 & 2 areas... (may take a moment)")

conserved_union <- conserved_gap12 |>
  summarise(geometry = st_union(geometry)) |>
  mutate(protected = "GAP 1 & 2")

gap12_acres <- sum(
  conserved_gap12$Gap1_acres + conserved_gap12$Gap2_acres,
  na.rm = TRUE
)
message(glue("GAP 1 & 2 coverage: {round(gap12_acres / 1e6, 2)} million acres"))
```

------------------------------------------------------------------------

## Classify Bee Cells by Protection Status

This chunk performs the core spatial join that every downstream analysis
depends on. Each 15 × 15 km bee grid cell is tested for intersection with the
dissolved GAP 1 & 2 polygon, and a logical `is_protected` column is added.

A left join is used so that cells with no overlap (i.e. entirely outside
protected lands) are retained with `NA` for the `protected` attribute, which
is then recoded to `FALSE`. This produces `bee_prot`, the single source of
protection-status truth used by both `gap-thresholds` and the sensitivity
analysis, eliminating the need for either chunk to recompute the intersection
independently.

```{r spatial-join}
bee_prot <- bee.data |>
  mutate(cell_id = row_number()) |>
  # Left join: every bee cell is kept; cells outside conserved areas get NA
  st_join(conserved_union["protected"], join = st_intersects, left = TRUE) |>
  # Guard against artefacts where one cell matches multiple polygons
  # (should not occur after dissolving, but handled defensively)
  group_by(cell_id) |>
  slice(1) |>
  ungroup() |>
  # A non-NA 'protected' value means the cell overlaps GAP 1 & 2 land
  mutate(is_protected = !is.na(protected))

n_prot   <- sum(bee_prot$is_protected)
n_unprot <- sum(!bee_prot$is_protected)

message(glue(
  "Protection status: {n_prot} cells overlap GAP 1 & 2 ",
  "({round(100 * n_prot / nrow(bee_prot), 1)}%), ",
  "{n_unprot} cells unprotected ",
  "({round(100 * n_unprot / nrow(bee_prot), 1)}%)"
))
```

------------------------------------------------------------------------

## Define Conservation Gap Thresholds

"High" richness and endemism are defined as cells at or above the **75th
percentile** of their respective distributions across all California grid cells.
This threshold captures the top quarter of biodiversity values while retaining
enough cells to be spatially meaningful. The quantile value is set via
`gap_quantile` and can be adjusted (e.g. to 0.90 for a more stringent
definition) without changing any other code.

The four priority categories assigned to each cell — high/lower richness or
endemism crossed with protected/unprotected — are used to colour all subsequent
maps.

```{r gap-thresholds}
# Adjust this value to change the definition of "high" biodiversity
gap_quantile <- 0.75

rich_thresh <- quantile(bee.data$richness, gap_quantile, na.rm = TRUE)
cwe_thresh  <- quantile(bee.data$cwe,      gap_quantile, na.rm = TRUE)

message(glue(
  "Richness threshold (≥{gap_quantile*100}th pctile): {round(rich_thresh, 1)} species\n",
  "CWE threshold     (≥{gap_quantile*100}th pctile): {round(cwe_thresh, 4)}"
))

# Build the gap-classified data frame from bee_prot (created in spatial-join)
bee_gaps <- bee_prot |>
  mutate(
    # Conservation gap flags: high diversity AND unprotected
    rich_gap = richness >= rich_thresh & !is_protected,
    cwe_gap  = cwe      >= cwe_thresh  & !is_protected,

    # Four-category richness × protection label for map colouring
    rich_priority = case_when(
      richness >= rich_thresh & !is_protected ~ "High richness, unprotected",
      richness >= rich_thresh &  is_protected ~ "High richness, protected",
      richness <  rich_thresh & !is_protected ~ "Lower richness, unprotected",
      TRUE                                    ~ "Lower richness, protected"
    ),

    # Four-category endemism × protection label for map colouring
    cwe_priority = case_when(
      cwe >= cwe_thresh & !is_protected ~ "High endemism, unprotected",
      cwe >= cwe_thresh &  is_protected ~ "High endemism, protected",
      cwe <  cwe_thresh & !is_protected ~ "Lower endemism, unprotected",
      TRUE                              ~ "Lower endemism, protected"
    )
  )

dual_gap_n <- sum(bee_gaps$rich_gap & bee_gaps$cwe_gap)
message(glue("Dual-gap cells (high richness AND endemism, unprotected): {dual_gap_n}"))
```

------------------------------------------------------------------------

## Sensitivity Analysis: Areal Overlap Thresholds

The binary protection classification above (`is_protected` = any intersection)
is the most inclusive possible definition. In practice, a bee grid cell might
overlap a protected area only at a corner, raising the question of whether it
should truly be counted as "protected." This section tests how gap counts
change when a minimum areal overlap threshold is required — specifically 0%
(any overlap, matching the main analysis), 25%, and 50%.

Because `bee_prot` is now available from the `spatial-join` chunk, the
sensitivity analysis can extend that object directly rather than recomputing
the intersection from scratch. The spatial intersection is run once to obtain
the overlap area of each cell with GAP 1 & 2 land; cells are then
reclassified under each threshold and gap counts are summarised.

The output is a single merged table formatted for direct inclusion in a
manuscript supplement.

```{r sensitivity-overlap-thresholds}
# ── Overlap thresholds to compare ─────────────────────────────────────────────
overlap_thresholds <- c(1.00, 0.75, 0.50, 0.25)

# Biodiversity thresholds mirror the main analysis (gap_quantile defined above)
rich_thresh_sens <- rich_thresh
cwe_thresh_sens  <- cwe_thresh

# ── Construct 15 km square polygon cells ──────────────────────────────────────
# The main analysis stores bee locations as points. For areal overlap we need
# polygons. st_buffer with endCapStyle = "SQUARE" produces axis-aligned squares.
bee_cells   <- bee_prot
cell_size_m <- 15000

geom_class <- class(sf::st_geometry(bee_cells))[[1]]

if (geom_class %in% c("sfc_POINT", "sfc_MULTIPOINT")) {
  message("bee_prot is point geometry — constructing 15 km square polygons.")
  sf::st_geometry(bee_cells) <- sf::st_buffer(
    sf::st_geometry(bee_cells),
    dist        = cell_size_m / 2,
    endCapStyle = "SQUARE"
  )
} else if (geom_class %in% c("sfc_POLYGON", "sfc_MULTIPOLYGON")) {
  message("bee_prot already has polygon geometry — using supplied polygons.")
} else {
  stop("Unsupported geometry type: ", geom_class)
}

sf::st_geometry(bee_cells) <- sf::st_make_valid(sf::st_geometry(bee_cells))
bee_cells$cell_area <- as.numeric(sf::st_area(bee_cells))

# ── Compute overlap area between each bee cell and the dissolved GAP 1/2 layer
# st_intersection clips each bee cell to the extent of the conserved union,
# producing a geometry whose area is the actual overlap.
bee_gap12_intersections <- suppressWarnings(suppressMessages(
  sf::st_intersection(
    bee_cells[, c("cell_id", "cell_area")],
    conserved_union["protected"]
  )
))

if (nrow(bee_gap12_intersections) > 0) {
  # Compute area outside mutate() to avoid pipe-placeholder issues with |>
  bee_gap12_intersections$overlap_area <- as.numeric(
    sf::st_area(bee_gap12_intersections)
  )
  overlap_by_cell <- bee_gap12_intersections |>
    sf::st_drop_geometry() |>
    dplyr::group_by(cell_id) |>
    dplyr::summarise(overlap_area = sum(overlap_area, na.rm = TRUE),
                     .groups = "drop")
} else {
  overlap_by_cell <- tibble::tibble(
    cell_id      = integer(),
    overlap_area = numeric()
  )
}

# Join overlap areas back to bee cells and compute the overlap proportion
bee_overlap <- bee_cells |>
  dplyr::left_join(overlap_by_cell, by = "cell_id") |>
  dplyr::mutate(
    overlap_area = tidyr::replace_na(overlap_area, 0),
    overlap_prop = pmin(overlap_area / cell_area, 1)  # cap at 1.0
  )

message(glue(
  "Overlap proportion — min: {round(min(bee_overlap$overlap_prop), 3)}, ",
  "median: {round(median(bee_overlap$overlap_prop), 3)}, ",
  "max: {round(max(bee_overlap$overlap_prop), 3)}"
))

# ── Reclassify protection and gap status under each threshold ──────────────────
# map_dfr iterates over each threshold value and returns a single stacked
# data frame with a column recording which threshold each row belongs to.
bee_overlap_long <- purrr::map_dfr(overlap_thresholds, function(thresh) {
  bee_overlap |>
    sf::st_drop_geometry() |>
    dplyr::mutate(
      overlap_threshold   = thresh,
      threshold_label     = scales::percent(thresh, accuracy = 1),
      is_protected_thresh = overlap_prop >= thresh,
      rich_gap_thresh     = richness >= rich_thresh_sens & !is_protected_thresh,
      cwe_gap_thresh      = cwe      >= cwe_thresh_sens  & !is_protected_thresh,
      dual_gap_thresh     = rich_gap_thresh & cwe_gap_thresh
    )
})

# ── Summary table 1: protection status by threshold ───────────────────────────
prot_summary_sensitivity <- bee_overlap_long |>
  dplyr::group_by(overlap_threshold, threshold_label) |>
  dplyr::summarise(
    n_cells_total       = dplyr::n(),
    n_cells_protected   = sum(is_protected_thresh, na.rm = TRUE),
    n_cells_unprotected = sum(!is_protected_thresh, na.rm = TRUE),
    pct_cells_protected = round(100 * mean(is_protected_thresh, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  dplyr::arrange(overlap_threshold)

# ── Summary table 2: gap counts by threshold ──────────────────────────────────
gap_summary_sensitivity <- bee_overlap_long |>
  dplyr::group_by(overlap_threshold, threshold_label) |>
  dplyr::summarise(
    rich_gap_cells = sum(rich_gap_thresh,  na.rm = TRUE),
    cwe_gap_cells  = sum(cwe_gap_thresh,   na.rm = TRUE),
    dual_gap_cells = sum(dual_gap_thresh,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(overlap_threshold)

# ── Combined publication table ─────────────────────────────────────────────────
# The two summaries are merged into a single wide table and formatted with
# kableExtra for manuscript submission. Column spanners distinguish cell-count
# columns from gap-count columns; a footnote provides the full methodological
# context needed to interpret the table without reference to the main text.
combined_sensitivity <- prot_summary_sensitivity |>
  dplyr::left_join(
    gap_summary_sensitivity,
    by = c("overlap_threshold", "threshold_label")
  ) |>
  dplyr::transmute(
    `Overlap threshold`  = threshold_label,
    `Total cells`        = n_cells_total,
    `Protected cells`    = n_cells_protected,
    `Unprotected cells`  = n_cells_unprotected,
    `% Protected`        = pct_cells_protected,
    `Richness gap cells` = rich_gap_cells,
    `Endemism gap cells` = cwe_gap_cells,
    `Dual-gap cells`     = dual_gap_cells
  )

knitr::kable(
  combined_sensitivity,
  caption = paste0(
    "Table S1. Sensitivity of conservation gap counts to minimum areal overlap ",
    "threshold with GAP Status 1 & 2 protected lands. ",
    "A cell is classified as 'protected' only if the stated proportion of its ",
    "15 × 15 km area overlaps with GAP 1 or 2 lands. ",
    "'Richness gap' = high species richness (≥ 75th percentile) and unprotected. ",
    "'Endemism gap' = high corrected weighted endemism (CWE; ≥ 75th percentile) ",
    "and unprotected. ",
    "'Dual-gap' = high in both richness and endemism and unprotected."
  ),
  align       = c("l", rep("r", 7)),
  format.args = list(big.mark = ",")
) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "bordered"),
    full_width        = TRUE,
    font_size         = 12
  ) |>
  kableExtra::add_header_above(
    c(" " = 1, "Cell counts" = 4, "Conservation gaps" = 3),
    bold = TRUE
  ) |>
  kableExtra::row_spec(0, bold = TRUE) |>
  kableExtra::footnote(
    general = paste0(
      "Bee grid cells are 15 × 15 km. ",
      "GAP 1 & 2 protected lands from CA Nature Terrestrial 30×30 Conserved Areas ",
      "(GreenInfo Network / USGS PAD-US, 2024). ",
      "Biodiversity thresholds: species richness ≥ ",
      round(rich_thresh_sens, 1), " spp and CWE ≥ ",
      round(cwe_thresh_sens, 4),
      " (75th percentile across all grid cells)."
    ),
    general_title     = "Note:",
    footnote_as_chunk = TRUE
  )

# Retain the spatial version for optional downstream GIS export
bee_overlap_sf <- bee_overlap |>
  dplyr::mutate(
    is_protected_0  = overlap_prop >= 0.00,
    is_protected_25 = overlap_prop >= 0.25,
    is_protected_50 = overlap_prop >= 0.50
  )

message(glue(
  "Created: bee_overlap, bee_overlap_long, bee_overlap_sf, ",
  "prot_summary_sensitivity, gap_summary_sensitivity, combined_sensitivity"
))

# Export sensitivity table as CSV for import into Word or Excel
write.csv(
  combined_sensitivity,
  file.path("..", "figures", "table_S1_sensitivity.csv"),
  row.names = FALSE
)

# Export as a formatted Word-ready table via flextable
# install.packages("flextable") if not already installed
library(flextable)

ft <- flextable(combined_sensitivity) |>
  set_caption("Table S1. Sensitivity of conservation gap counts to minimum areal overlap threshold") |>
  add_header_row(
    values = c("", "Cell counts", "Conservation gaps"),
    colwidths = c(1, 4, 3)
  ) |>
  bold(part = "header") |>
  align(align = "right", part = "body") |>
  align(j = 1, align = "left", part = "body") |>
  autofit()

save_as_docx(ft, path = file.path("..", "figures", "table_S1_sensitivity.docx"))
```

------------------------------------------------------------------------

## Maps

### Shared map elements

Before drawing individual maps, this chunk defines three reusable objects that
are composed into every panel:

- `map_theme` — a publication-ready ggplot theme (white background, clean
  typography, bottom legend) that conforms to most ecology journal figure
  standards.
- `ca_outline_layer` — a dark grey California state boundary, downloaded from
  the Census Bureau via `tigris` and reprojected to CA Albers. A convex hull
  of the conserved areas is used as a fallback if `tigris` fails.
- `conserved_base` — a light grey-green fill layer showing GAP 1 & 2 parcels,
  drawn beneath all data layers so the conservation footprint provides context
  without obscuring bee diversity values.

All colours are chosen to be distinguishable under common forms of
colour-vision deficiency and to reproduce clearly in greyscale, following the
Wong (2011) palette.

```{r map-theme}
bg_col       <- "#ffffff"
text_col     <- "#1a1a1a"
subtitle_col <- "#555555"
border_col   <- "#333333"

map_theme <- theme_void(base_size = 11) +
  theme(
    plot.background  = element_rect(fill = bg_col, color = NA),
    panel.background = element_rect(fill = bg_col, color = NA),
    plot.title    = element_text(color = text_col, face = "bold",
                                 size = 12, margin = margin(b = 4)),
    plot.subtitle = element_text(color = subtitle_col, size = 9,
                                 margin = margin(b = 8)),
    legend.position    = "bottom",
    legend.direction   = "horizontal",
    legend.title       = element_text(color = text_col, size = 9, face = "bold"),
    legend.text        = element_text(color = text_col, size = 8),
    legend.background  = element_rect(fill = bg_col, color = NA),
    legend.key         = element_rect(fill = bg_col, color = NA),
    legend.key.width   = unit(1.5, "cm"),
    legend.key.height  = unit(0.35, "cm"),
    plot.margin = margin(6, 8, 6, 8)
  )

ca_outline_layer <- tryCatch({
  ca_state <- tigris::states(cb = TRUE, progress_bar = FALSE) |>
    dplyr::filter(STUSPS == "CA") |>
    sf::st_transform(crs = 3310)
  list(
    geom_sf(data = ca_state, fill = NA, color = border_col, linewidth = 0.4),
    coord_sf(expand = FALSE)
  )
}, error = function(e) {
  ca_hull <- sf::st_union(conserved |> sf::st_make_valid()) |>
    sf::st_convex_hull()
  list(
    geom_sf(data = ca_hull, fill = NA, color = border_col, linewidth = 0.4),
    coord_sf(expand = FALSE)
  )
})

# Light warm grey: neutral "land cover" feel, does not compete with data scales
conserved_base <- geom_sf(
  data  = conserved_gap12,
  fill  = "#c8cfc8",
  color = NA,
  alpha = 0.80
)
```

------------------------------------------------------------------------

### Map 1 — Bee Species Richness Overlaid on Conserved Lands

Map 1 shows the raw species richness of each grid cell, coloured using the
viridis palette (purple = low, yellow = high). GAP 1 & 2 conserved lands
appear as a light grey background layer. This map communicates the overall
spatial pattern of bee diversity across California before any gap
classification is applied.

```{r map-richness-overlay, fig.alt="Bee richness across California with GAP 1 & 2 conserved areas shown in gray"}
map1 <- ggplot() +
  conserved_base +
  ca_outline_layer +
  geom_sf(
    data  = bee.data,
    aes(color = richness),
    size  = 1.0,
    alpha = 0.85,
    shape = 16
  ) +
  scale_color_viridis_c(
    option    = "viridis",
    direction = 1,
    name      = "Species richness",
    labels    = label_number(accuracy = 1),
    guide     = guide_colorbar(title.position = "top", title.hjust = 0.5,
                               barwidth = 12, barheight = 0.4)
  ) +
  labs(
    title    = "Bee Species Richness Across California",
    subtitle = "Light grey = GAP 1 & 2 conserved lands  ·  Points = bee survey grid cells"
  ) +
  map_theme

map1
```

------------------------------------------------------------------------

### Map 2 — Bee Endemism (CWE) Overlaid on Conserved Lands

Map 2 shows corrected weighted endemism (CWE), a metric that upweights
species with restricted geographic ranges. A high CWE value indicates that a
grid cell harbours a disproportionate share of range-restricted bee species —
the kind of diversity that cannot be recovered if a site is lost. The mako
palette (distinct from viridis on Map 1) makes the two maps visually
distinguishable in multi-panel figures.

```{r map-cwe-overlay, fig.alt="Bee endemism (CWE) across California with GAP 1 & 2 conserved areas shown in gray"}
map2 <- ggplot() +
  conserved_base +
  ca_outline_layer +
  geom_sf(
    data  = bee.data,
    aes(color = cwe),
    size  = 1.0,
    alpha = 0.85,
    shape = 16
  ) +
  scale_color_viridis_c(
    option    = "mako",
    direction = 1,
    name      = "Bee rarity (CWE)",
    labels    = label_number(accuracy = 0.001),
    guide     = guide_colorbar(title.position = "top", title.hjust = 0.5,
                               barwidth = 12, barheight = 0.4)
  ) +
  labs(
    title    = "Bee Rarity (Community-Weighted Endemism) Across California",
    subtitle = "Light grey = GAP 1 & 2 conserved lands  ·  CWE highlights range-restricted species"
  ) +
  map_theme

map2
```

------------------------------------------------------------------------

### Map 3 — Conservation Gap: High Richness, Unprotected

Map 3 is the first gap map. Cells are divided into four categories based on
whether their richness exceeds the 75th-percentile threshold and whether they
fall inside or outside GAP 1 & 2 lands. The key category — high richness AND
unprotected (orange/vermillion) — represents locations where the conservation
network fails to capture above-average bee diversity. Gap cells are rendered
on top of non-gap cells and at larger point size to ensure they are legible at
publication resolution.

```{r map-richness-gap, fig.alt="Conservation gap map showing high bee richness areas outside GAP 1 & 2 lands"}
# Vermillion for the gap (Wong 2011 colorblind-safe palette),
# blue for protected high-richness cells, greys for background
gap_colors_rich <- c(
  "High richness, unprotected"   = "#D55E00",
  "High richness, protected"     = "#0072B2",
  "Lower richness, unprotected"  = "#BBBBBB",
  "Lower richness, protected"    = "#888888"
)

map3 <- ggplot() +
  conserved_base +
  ca_outline_layer +
  # Render lower-priority cells first so gap cells sit on top
  geom_sf(
    data  = bee_gaps |> filter(!rich_gap),
    aes(color = rich_priority),
    size  = 1.0, alpha = 0.65, shape = 16
  ) +
  geom_sf(
    data  = bee_gaps |> filter(rich_gap),
    aes(color = rich_priority),
    size  = 1.5, alpha = 1.0, shape = 16
  ) +
  scale_color_manual(
    values = gap_colors_rich,
    name   = NULL,
    guide  = guide_legend(
      override.aes   = list(size = 3),
      nrow           = 2,
      title.position = "top"
    )
  ) +
  labs(
    title    = "Conservation Gap: Bee Species Richness",
    subtitle = glue(
      "Orange = high richness (≥ {round(rich_thresh,1)} spp) AND outside protected lands  ·  ",
      "Threshold = {gap_quantile*100}th percentile"
    )
  ) +
  map_theme

map3
```

------------------------------------------------------------------------

### Map 4 — Conservation Gap: High Endemism, Unprotected

Map 4 mirrors Map 3 but uses CWE as the biodiversity metric. Because CWE
emphasises range-restricted species rather than total counts, the spatial
pattern of endemism gaps may differ substantially from richness gaps — areas
with moderate total richness can still harbour many narrow-range endemics.
Using the same colour scheme as Map 3 allows direct visual comparison across
the two gap dimensions.

```{r map-cwe-gap, fig.alt="Conservation gap map showing high bee endemism areas outside GAP 1 & 2 lands"}
gap_colors_cwe <- c(
  "High endemism, unprotected"   = "#D55E00",
  "High endemism, protected"     = "#0072B2",
  "Lower endemism, unprotected"  = "#BBBBBB",
  "Lower endemism, protected"    = "#888888"
)

map4 <- ggplot() +
  conserved_base +
  ca_outline_layer +
  geom_sf(
    data  = bee_gaps |> filter(!cwe_gap),
    aes(color = cwe_priority),
    size  = 1.0, alpha = 0.65, shape = 16
  ) +
  geom_sf(
    data  = bee_gaps |> filter(cwe_gap),
    aes(color = cwe_priority),
    size  = 1.5, alpha = 1.0, shape = 16
  ) +
  scale_color_manual(
    values = gap_colors_cwe,
    name   = NULL,
    guide  = guide_legend(
      override.aes   = list(size = 3),
      nrow           = 2,
      title.position = "top"
    )
  ) +
  labs(
    title    = "Conservation Gap: Bee Rarity (CWE)",
    subtitle = glue(
      "Orange = high rarity (≥ {round(cwe_thresh,4)} CWE) AND outside protected lands  ·  ",
      "Threshold = {gap_quantile*100}th percentile"
    )
  ) +
  map_theme

map4
```

------------------------------------------------------------------------

### Map 5 — Dual-Gap Priority Cells

Map 5 identifies the highest-priority conservation gaps: cells that are
simultaneously above the richness threshold AND above the endemism threshold
AND outside GAP 1 & 2 lands. These dual-gap cells represent locations where
protecting new land would advance two independent conservation objectives at
once. A five-category palette distinguishes dual-gap cells (red) from
single-dimension gaps (amber for richness only, sky blue for endemism only),
already-protected high-diversity cells (teal), and background cells (light
grey).

```{r map-dual-gap, fig.alt="Map showing cells high in both bee richness and endemism but outside GAP 1 & 2 lands"}
bee_gaps <- bee_gaps |>
  mutate(
    dual_gap = rich_gap & cwe_gap,
    dual_priority = case_when(
      dual_gap      ~ "High richness + endemism, unprotected",
      rich_gap      ~ "High richness only, unprotected",
      cwe_gap       ~ "High endemism only, unprotected",
      is_protected  ~ "Protected (GAP 1 & 2)",
      TRUE          ~ "Lower diversity, unprotected"
    )
  )

n_dual <- sum(bee_gaps$dual_gap, na.rm = TRUE)
message(glue(
  "{n_dual} cells are high in BOTH richness AND endemism and are unprotected — ",
  "these are the highest-priority dual conservation gaps."
))

dual_colors <- c(
  "High richness + endemism, unprotected" = "#CC0000",
  "High richness only, unprotected"       = "#E69F00",
  "High endemism only, unprotected"       = "#56B4E9",
  "Protected (GAP 1 & 2)"                = "#009E73",
  "Lower diversity, unprotected"          = "#DDDDDD"
)

map5 <- ggplot() +
  conserved_base +
  ca_outline_layer +
  # Render from lowest to highest priority so key cells are never occluded
  geom_sf(
    data  = bee_gaps |> filter(!rich_gap & !cwe_gap),
    aes(color = dual_priority),
    size  = 0.8, alpha = 0.5, shape = 16
  ) +
  geom_sf(
    data  = bee_gaps |> filter(rich_gap | cwe_gap) |> filter(!dual_gap),
    aes(color = dual_priority),
    size  = 1.2, alpha = 0.8, shape = 16
  ) +
  geom_sf(
    data  = bee_gaps |> filter(dual_gap),
    aes(color = dual_priority),
    size  = 2.0, alpha = 1.0, shape = 16
  ) +
  scale_color_manual(
    values = dual_colors,
    name   = NULL,
    guide  = guide_legend(
      override.aes   = list(size = 3),
      nrow           = 3,
      title.position = "top"
    )
  ) +
  labs(
    title    = "Dual Conservation Gap: High Bee Richness AND Endemism, Unprotected",
    subtitle = glue(
      "Red = {n_dual} cells unprotected AND high in both richness ",
      "(≥{round(rich_thresh,1)} spp) AND endemism (CWE ≥{round(cwe_thresh,4)})  ·  ",
      "Threshold = {gap_quantile*100}th percentile"
    )
  ) +
  map_theme

map5
```

------------------------------------------------------------------------

## Summary Statistics

The three tables below quantify the gap patterns shown in Maps 3–5. Because
these statistics are computed from the same `bee_gaps` object used to generate
the maps, any change to `gap_quantile` in the `gap-thresholds` chunk
automatically propagates through both the maps and the summary tables.

### Richness gap

```{r summary-richness}
bee_gaps |>
  st_drop_geometry() |>
  group_by(rich_priority) |>
  summarise(
    n_cells         = n(),
    mean_richness   = round(mean(richness, na.rm = TRUE), 1),
    median_richness = round(median(richness, na.rm = TRUE), 1)
  ) |>
  arrange(desc(mean_richness)) |>
  knitr::kable(caption = "Bee richness by conservation status")
```

### Endemism gap

```{r summary-cwe}
bee_gaps |>
  st_drop_geometry() |>
  group_by(cwe_priority) |>
  summarise(
    n_cells    = n(),
    mean_cwe   = round(mean(cwe, na.rm = TRUE), 4),
    median_cwe = round(median(cwe, na.rm = TRUE), 4)
  ) |>
  arrange(desc(mean_cwe)) |>
  knitr::kable(caption = "Bee endemism (CWE) by conservation status")
```

### Overlap: cells that are gaps for both richness and endemism

```{r summary-overlap}
dual_gap <- bee_gaps |>
  st_drop_geometry() |>
  filter(rich_gap & cwe_gap)

# cat() is used instead of message() so this count is never suppressed
# by execute: message: false in the YAML header
cat(glue(
  "{nrow(dual_gap)} grid cells are high in BOTH richness AND endemism ",
  "and fall outside GAP 1 & 2 protected areas — these are the highest-priority gaps.\n"
))

# A visible summary table ensures this chunk always produces rendered output,
# preventing Quarto from skipping it in sessions where message() is suppressed
dual_gap |>
  summarise(
    n_dual_gap_cells    = n(),
    mean_richness       = round(mean(richness, na.rm = TRUE), 1),
    mean_cwe            = round(mean(cwe, na.rm = TRUE), 4)
  ) |>
  knitr::kable(
    caption   = "Dual-gap cells: high in both richness and endemism, unprotected",
    col.names = c("Dual-gap cells (n)", "Mean richness", "Mean CWE")
  )
```

------------------------------------------------------------------------

## Export Gap Layers

The classified grid is exported as two GeoPackage files for use in QGIS or
ArcGIS. The full classified grid (`bee_conservation_gaps.gpkg`) retains all
cells with their gap-status columns. The priority subset
(`bee_priority_gaps.gpkg`) contains only cells that are a gap on at least one
biodiversity dimension; this smaller file loads and renders faster in GIS
software and is suitable for direct use in planning workflows.

```{r export}
out_dir <- "outputs"
dir.create(out_dir, showWarnings = FALSE)

st_write(
  bee_gaps,
  file.path(out_dir, "bee_conservation_gaps.gpkg"),
  append = FALSE
)

priority_gaps <- bee_gaps |>
  filter(rich_gap | cwe_gap)

st_write(
  priority_gaps,
  file.path(out_dir, "bee_priority_gaps.gpkg"),
  append = FALSE
)

message(glue(
  "Exported {nrow(bee_gaps)} classified cells and ",
  "{nrow(priority_gaps)} priority gap cells to {out_dir}/"
))
```

------------------------------------------------------------------------

## Export Manuscript Figures

Maps 3, 4, and 5 (the gap maps) are exported as 600 dpi LZW-compressed TIFF
files at 174 × 160 mm — a standard two-column figure size for journals such
as *iScience*, *Oecologia*, and *Diversity and Distributions*. Maps 1 and 2
(the overlay maps) are supplementary and can be added to this list by
extending `fig_specs` if needed.

```{r export-figures}
fig_dir <- file.path("..", "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

fig_specs <- list(
  list(plot = map3, name = "fig3_richness_gap"),
  list(plot = map4, name = "fig4_cwe_gap"),
  list(plot = map5, name = "fig5_dual_gap")
)

for (fig in fig_specs) {
  ggsave(
    filename    = file.path(fig_dir, paste0(fig$name, ".tiff")),
    plot        = fig$plot,
    width       = 174,
    height      = 160,
    units       = "mm",
    dpi         = 600,
    compression = "lzw"
  )
  message(glue("Saved: {fig$name}.tiff"))
}

message(glue("All figures exported to: {normalizePath(fig_dir)}"))
```

------------------------------------------------------------------------

## Data Integrity Validation

This section runs after all analysis chunks have executed. It checks each
of the thirteen issues identified in the crosscheck review: silent data
loss, NA propagation, row-count changes across joins, type mismatches,
geometry validity, CRS consistency, and output visibility. Results are
collected into a single table so that all issues are visible at once
rather than stopping on the first failure.

Each check is assigned one of three statuses:

- **FAIL** — a problem that will produce incorrect results and must be
  resolved before any findings are reported.
- **WARN** — a condition that may be acceptable but requires a conscious
  decision; document your reasoning if you proceed.
- **PASS** — the check found no problem.

The table is rendered twice: as plain text to the console (visible when
running interactively) and as a formatted HTML table in the rendered
document. Note that because `execute: message: false` is set in the YAML
header, `message()` calls elsewhere in the script are suppressed in the
HTML output; this chunk uses `cat()` for the console summary so it is
always visible regardless of that setting.

```{r validation, message=TRUE}
# ==============================================================================
# DATA INTEGRITY VALIDATION
# Run this chunk after all preceding chunks have executed successfully.
# Each check corresponds to a numbered item in the crosscheck list.
# Results are collected into a summary table printed at the end.
# A script that passes all checks prints only PASS lines.
# Any FAIL or WARN line requires investigation before results are reported.
# ==============================================================================

# ── Helper: collect results rather than stop on first failure ─────────────────
# Each check appends one row to `checks`: severity, id, description, detail.
checks <- tibble::tibble(
  severity    = character(),
  id          = character(),
  description = character(),
  detail      = character()
)

flag <- function(severity, id, description, detail = "") {
  checks <<- dplyr::bind_rows(
    checks,
    tibble::tibble(
      severity    = severity,
      id          = id,
      description = description,
      detail      = as.character(detail)
    )
  )
}

pass <- function(id, description, detail = "") flag("PASS", id, description, detail)
warn <- function(id, description, detail = "") flag("WARN", id, description, detail)
fail <- function(id, description, detail = "") flag("FAIL", id, description, detail)

# ==============================================================================
# CHECK 1 — Rows dropped by drop_na() on coordinates (CSV path)
# We cannot rerun the CSV path here, but we can verify that no NA coordinates
# survived into bee.data after loading.
# ==============================================================================
{
  n_na_coords <- if (inherits(sf::st_geometry(bee.data), "sfc_POINT")) {
    coords <- sf::st_coordinates(bee.data)
    sum(is.na(coords[, 1]) | is.na(coords[, 2]))
  } else {
    0L  # polygon geometry: coordinate NAs not directly checkable this way
  }

  if (n_na_coords > 0) {
    fail("1-coord-na",
         "NA coordinates present in bee.data after loading",
         glue::glue("{n_na_coords} cells have NA x or y — these were or will be dropped"))
  } else {
    pass("1-coord-na",
         "No NA coordinates in bee.data",
         glue::glue("{nrow(bee.data)} cells loaded"))
  }
}

# ==============================================================================
# CHECK 2 — NA values in richness and cwe
# NA richness/cwe causes case_when() to silently assign cells to the catch-all
# category ("Lower richness, protected") regardless of actual protection status.
# ==============================================================================
{
  n_na_rich <- sum(is.na(bee_prot$richness))
  n_na_cwe  <- sum(is.na(bee_prot$cwe))

  if (n_na_rich > 0) {
    fail("2-na-richness",
         "NA values in richness column of bee_prot",
         glue::glue("{n_na_rich} cells — silently misclassified by case_when()"))
  } else {
    pass("2-na-richness", "No NA values in richness",
         glue::glue("n = {nrow(bee_prot)}"))
  }

  if (n_na_cwe > 0) {
    fail("2-na-cwe",
         "NA values in cwe column of bee_prot",
         glue::glue("{n_na_cwe} cells — silently misclassified by case_when()"))
  } else {
    pass("2-na-cwe", "No NA values in cwe",
         glue::glue("n = {nrow(bee_prot)}"))
  }

  # Secondary: confirm gap flags contain no NAs (NA propagation reached bee_gaps)
  n_na_rich_gap <- sum(is.na(bee_gaps$rich_gap))
  n_na_cwe_gap  <- sum(is.na(bee_gaps$cwe_gap))

  if (n_na_rich_gap > 0 || n_na_cwe_gap > 0) {
    fail("2-na-gap-flags",
         "NA values in rich_gap or cwe_gap in bee_gaps",
         glue::glue("rich_gap NAs: {n_na_rich_gap}, cwe_gap NAs: {n_na_cwe_gap}"))
  } else {
    pass("2-na-gap-flags", "No NA values in rich_gap or cwe_gap")
  }
}

# ==============================================================================
# CHECK 3 — Row count preserved by spatial-join deduplication
# bee_prot should have exactly the same number of rows as bee.data.
# Any excess indicates that slice(1) silently discarded duplicate matches.
# ==============================================================================
{
  n_bee_data <- nrow(bee.data)
  n_bee_prot <- nrow(bee_prot)

  if (n_bee_prot != n_bee_data) {
    fail("3-join-dedup",
         "Row count changed after spatial-join deduplication",
         glue::glue("bee.data: {n_bee_data} rows, bee_prot: {n_bee_prot} rows — ",
                    "difference: {n_bee_prot - n_bee_data}"))
  } else {
    pass("3-join-dedup",
         "Row count unchanged after spatial join",
         glue::glue("{n_bee_prot} rows in bee_prot == {n_bee_data} rows in bee.data"))
  }
}

# ==============================================================================
# CHECK 4 — Gap1_acres and Gap2_acres column types in conserved
# If stored as character, the filter (> 0) evaluates to NA, not FALSE, and
# conserved_gap12 may be empty without any error.
# ==============================================================================
{
  cls_gap1 <- class(conserved$Gap1_acres)
  cls_gap2 <- class(conserved$Gap2_acres)
  numeric_classes <- c("numeric", "integer", "double")

  if (!cls_gap1 %in% numeric_classes) {
    fail("4-acres-type",
         "Gap1_acres is not numeric — filter(Gap1_acres > 0) may have failed silently",
         glue::glue("class: {cls_gap1}"))
  } else if (!cls_gap2 %in% numeric_classes) {
    fail("4-acres-type",
         "Gap2_acres is not numeric — filter(Gap2_acres > 0) may have failed silently",
         glue::glue("class: {cls_gap2}"))
  } else {
    pass("4-acres-type",
         "Gap1_acres and Gap2_acres are numeric",
         glue::glue("classes: {cls_gap1}, {cls_gap2}"))
  }

  # Secondary: conserved_gap12 must be non-empty
  if (nrow(conserved_gap12) == 0) {
    fail("4-gap12-empty",
         "conserved_gap12 is empty — no GAP 1 or 2 polygons found",
         "Check column names and types in the 30x30 shapefile")
  } else {
    pass("4-gap12-empty",
         "conserved_gap12 is non-empty",
         glue::glue("{nrow(conserved_gap12)} polygons with GAP 1 or 2 acreage"))
  }
}

# ==============================================================================
# CHECK 5 — Zero or degenerate cell areas in sensitivity analysis
# overlap_area is computed by calling sf::st_area() directly on
# bee_gap12_intersections (assigned via $<- before the summarise pipe).
# If any bee_cells polygon is degenerate (area = 0), dividing overlap_area
# by cell_area produces NaN, which sum(..., na.rm = TRUE) silently excludes
# from gap counts — understating the number of unprotected cells.
# ==============================================================================
{
  if (exists("bee_cells") && "cell_area" %in% names(bee_cells)) {
    n_zero_area <- sum(bee_cells$cell_area == 0, na.rm = TRUE)
    n_na_area   <- sum(is.na(bee_cells$cell_area))

    if (n_zero_area > 0) {
      fail("5-zero-area",
           "Degenerate polygons with cell_area == 0 in bee_cells",
           glue::glue("{n_zero_area} cells — overlap_prop will be NaN for these"))
    } else {
      pass("5-zero-area", "No zero-area cells in bee_cells")
    }

    if (n_na_area > 0) {
      fail("5-na-area",
           "NA values in cell_area in bee_cells",
           glue::glue("{n_na_area} cells"))
    } else {
      pass("5-na-area", "No NA values in cell_area in bee_cells")
    }

    # Check for NaN in overlap_prop — the downstream symptom of a zero cell_area.
    # st_area() is called on bee_gap12_intersections directly (not inside mutate),
    # so the area values themselves should always be non-negative numerics;
    # NaN here means cell_area was zero, not that st_area() failed.
    if (exists("bee_overlap") && "overlap_prop" %in% names(bee_overlap)) {
      n_nan_prop <- sum(is.nan(sf::st_drop_geometry(bee_overlap)$overlap_prop))
      if (n_nan_prop > 0) {
        fail("5-nan-overlap-prop",
             "NaN values in overlap_prop in bee_overlap",
             glue::glue("{n_nan_prop} cells — caused by division by zero cell_area"))
      } else {
        pass("5-nan-overlap-prop", "No NaN values in overlap_prop")
      }
    }
  } else {
    warn("5-zero-area",
         "bee_cells or cell_area not found — sensitivity chunk may not have run",
         "Skipping check 5")
  }
}

# ==============================================================================
# CHECK 6 — CRS of bee.data and conserved_union
# Both must be EPSG 3310 (CA Albers) for the spatial join to be valid.
# ==============================================================================
{
  epsg_bee <- sf::st_crs(bee.data)$epsg

  if (is.na(epsg_bee)) {
    warn("6-crs-bee",
         "bee.data CRS has no EPSG code — cannot confirm EPSG 3310",
         glue::glue("CRS string: {sf::st_crs(bee.data)$input}"))
  } else if (epsg_bee != 3310L) {
    fail("6-crs-bee",
         "bee.data is not in CA Albers (EPSG 3310) after st_transform",
         glue::glue("Detected EPSG: {epsg_bee}"))
  } else {
    pass("6-crs-bee", "bee.data CRS confirmed as EPSG 3310 (CA Albers)")
  }

  epsg_union <- sf::st_crs(conserved_union)$epsg
  if (!isTRUE(epsg_union == 3310L)) {
    fail("6-crs-union",
         "conserved_union is not in EPSG 3310",
         glue::glue("Detected EPSG: {epsg_union}"))
  } else {
    pass("6-crs-union", "conserved_union CRS confirmed as EPSG 3310")
  }
}

# ==============================================================================
# CHECK 7 — conserved_union geometry is valid and non-empty
# An empty or invalid dissolve means every bee cell gets is_protected = FALSE.
# ==============================================================================
{
  union_empty <- sf::st_is_empty(conserved_union$geometry)
  union_valid <- sf::st_is_valid(conserved_union$geometry)

  if (any(union_empty)) {
    fail("7-union-empty",
         "conserved_union geometry is empty — all cells will appear unprotected",
         "st_union() may have failed on invalid input geometries")
  } else {
    pass("7-union-empty", "conserved_union geometry is non-empty")
  }

  if (!all(union_valid)) {
    fail("7-union-valid",
         "conserved_union geometry is invalid",
         "Run st_make_valid(conserved_union) and re-dissolve")
  } else {
    pass("7-union-valid", "conserved_union geometry is valid")
  }
}

# ==============================================================================
# CHECK 8 — Row count preserved by left_join in sensitivity chunk
# bee_overlap must have exactly as many rows as bee_cells.
# Extra rows mean overlap_by_cell had duplicate cell_ids, inflating gap counts.
# ==============================================================================
{
  if (exists("bee_overlap") && exists("bee_cells")) {
    n_cells   <- nrow(bee_cells)
    n_overlap <- nrow(bee_overlap)

    if (n_overlap != n_cells) {
      fail("8-overlap-rowcount",
           "Row count changed after left_join in sensitivity chunk",
           glue::glue("bee_cells: {n_cells} rows, bee_overlap: {n_overlap} rows — ",
                      "overlap_by_cell may contain duplicate cell_ids"))
    } else {
      pass("8-overlap-rowcount",
           "Row count unchanged after sensitivity left_join",
           glue::glue("{n_overlap} rows"))
    }
  } else {
    warn("8-overlap-rowcount",
         "bee_overlap or bee_cells not found — sensitivity chunk may not have run",
         "Skipping check 8")
  }
}

# ==============================================================================
# CHECK 9 — Sensitivity 0% threshold matches main analysis dual-gap count
# At 0% overlap, every intersecting cell is "protected" — identical to the
# is_protected flag from spatial-join. The dual-gap count must therefore
# equal dual_gap_n. A mismatch signals an artefact between the two pipelines.
# This was confirmed 8 april 2026 - Gretchen LeBuhn
# ==============================================================================
{
  if (exists("gap_summary_sensitivity") && exists("dual_gap_n")) {
    sens_dual_0 <- gap_summary_sensitivity |>
      dplyr::filter(overlap_threshold == 0) |>
      dplyr::pull(dual_gap_cells)

    if (length(sens_dual_0) == 0) {
      warn("9-sens-consistency",
           "Could not find 0% threshold row in gap_summary_sensitivity",
           "Confirmed 8 apr 2026.  So code changed to not include 0%")
    } else if (sens_dual_0 != dual_gap_n) {
      fail("9-sens-consistency",
           "Dual-gap count at 0% overlap threshold does not match main analysis",
           glue::glue("Main analysis dual_gap_n: {dual_gap_n}, ",
                      "Sensitivity at 0%: {sens_dual_0} — ",
                      "investigate geometry or join differences between pipelines"))
    } else {
      pass("9-sens-consistency",
           "Sensitivity 0% dual-gap count matches main analysis",
           glue::glue("Both = {dual_gap_n}"))
    }
  } else {
    warn("9-sens-consistency",
         "gap_summary_sensitivity or dual_gap_n not found",
         "Skipping check 9 — ensure sensitivity chunk ran after gap-thresholds")
  }
}

# ==============================================================================
# CHECK 10 — YAML execute:message:false suppresses diagnostic output
# This cannot be tested in R, but is flagged as a reminder so it appears in
# the validation output and is not overlooked before submission.
# ==============================================================================
{
  warn("10-message-suppressed",
       "YAML execute: message: false suppresses all message() output in rendered HTML",
       paste0("Verify counts manually before submission, or use message: true ",
              "during review and cat() for counts that must appear in the final document"))
}

# ==============================================================================
# CHECK 11 — Multiple shapefiles in the conserved areas zip
# If > 1 .shp was found, conserved_path[1] was used silently.
# ==============================================================================
{
  if (exists("conserved_dir") && dir.exists(conserved_dir)) {
    all_shp <- list.files(
      conserved_dir,
      pattern     = "\\.shp$",
      full.names  = TRUE,
      recursive   = TRUE,
      ignore.case = TRUE
    )
    if (length(all_shp) > 1) {
      warn("11-multi-shp",
           "Multiple .shp files found in zip — only the first was read",
           glue::glue("Files: {paste(basename(all_shp), collapse = ', ')}"))
    } else if (length(all_shp) == 1) {
      pass("11-multi-shp",
           "Exactly one .shp file in conserved areas zip",
           glue::glue("File: {basename(all_shp)}"))
    } else {
      warn("11-multi-shp",
           "No .shp files found when rechecking conserved_dir",
           "Temp directory may have been cleaned; verify manually")
    }
  } else {
    warn("11-multi-shp",
         "Temp directory conserved_dir no longer exists — cannot recheck zip contents",
         "Acceptable if tempfiles were cleaned; verify manually if in doubt")
  }
}

# ==============================================================================
# CHECK 12 — Double-counting of acres where both Gap1_acres and Gap2_acres > 0
# Affects only the summary message, not the analysis, but flags for awareness.
# ==============================================================================
{
  n_both <- sum(
    conserved_gap12$Gap1_acres > 0 & conserved_gap12$Gap2_acres > 0,
    na.rm = TRUE
  )
  if (n_both > 0) {
    warn("12-acres-double",
         "Parcels with both Gap1_acres > 0 and Gap2_acres > 0 exist",
         glue::glue("{n_both} parcels — the GAP 1 & 2 acreage summary message ",
                    "may overstate protected area. Analysis unaffected."))
  } else {
    pass("12-acres-double",
         "No parcels with both Gap1_acres and Gap2_acres > 0",
         "Acreage summary message is not double-counting")
  }
}

# ==============================================================================
# CHECK 13 — dual_gap object exists and count matches gap-thresholds
# The summary-overlap chunk creates dual_gap and reports its count via cat()
# (not message(), which is suppressed by execute: message: false) and renders
# a knitr::kable() to ensure the chunk always produces visible output.
# A mismatch in counts signals that bee_gaps was modified between chunks.
# ==============================================================================
{
  if (!exists("dual_gap")) {
    fail("13-dual-gap-exists",
         "Object 'dual_gap' not found — summary-overlap chunk did not run",
         paste0("The chunk now uses cat() and knitr::kable() to force visible output. ",
                "If dual_gap is still missing, check for an error in that chunk."))
  } else {
    n_dual_overlap <- nrow(dual_gap)
    if (n_dual_overlap != dual_gap_n) {
      fail("13-dual-gap-count",
           "dual_gap row count does not match dual_gap_n from gap-thresholds",
           glue::glue("dual_gap rows: {n_dual_overlap}, dual_gap_n: {dual_gap_n} — ",
                      "bee_gaps may have been modified between the two chunks"))
    } else {
      pass("13-dual-gap-count",
           "dual_gap exists and count matches gap-thresholds",
           glue::glue("n = {n_dual_overlap}"))
    }
  }
}

# ==============================================================================
# SUMMARY TABLE
# ==============================================================================

# ── Console output (always visible, not affected by message: false) ────────────
cat("\n")
cat("=======================================================================\n")
cat("  DATA INTEGRITY VALIDATION SUMMARY\n")
cat("=======================================================================\n\n")

n_fail <- sum(checks$severity == "FAIL")
n_warn <- sum(checks$severity == "WARN")
n_pass <- sum(checks$severity == "PASS")

ordered_checks <- dplyr::bind_rows(
  dplyr::filter(checks, severity == "FAIL"),
  dplyr::filter(checks, severity == "WARN"),
  dplyr::filter(checks, severity == "PASS")
)

for (i in seq_len(nrow(ordered_checks))) {
  row <- ordered_checks[i, ]
  prefix <- switch(row$severity,
    "FAIL" = "[ FAIL ]",
    "WARN" = "[ WARN ]",
    "PASS" = "[ pass ]"
  )
  cat(sprintf("%-10s  %s — %s\n", prefix, row$id, row$description))
  if (nchar(row$detail) > 0) {
    detail_wrapped <- strwrap(row$detail, width = 68,
                              prefix = "             ")
    cat(paste(detail_wrapped, collapse = "\n"), "\n")
  }
  cat("\n")
}

cat("-----------------------------------------------------------------------\n")
cat(sprintf("  %d FAIL   %d WARN   %d pass   (%d checks total)\n\n",
            n_fail, n_warn, n_pass, nrow(checks)))

if (n_fail > 0) {
  cat("  ACTION REQUIRED: resolve all FAIL items before reporting results.\n")
} else if (n_warn > 0) {
  cat("  Review WARN items and confirm they are acceptable before submission.\n")
} else {
  cat("  All checks passed. Proceed to submission.\n")
}
cat("=======================================================================\n\n")

# ── HTML table (rendered document) ────────────────────────────────────────────
# Row indices for conditional colouring must be computed on the ordered frame
fail_rows <- which(ordered_checks$severity == "FAIL")
warn_rows <- which(ordered_checks$severity == "WARN")

ordered_checks |>
  dplyr::select(severity, id, description, detail) |>
  knitr::kable(
    caption   = "Data integrity validation results",
    col.names = c("Status", "Check ID", "Description", "Detail"),
    align     = c("c", "l", "l", "l")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "condensed", "bordered"),
    full_width        = TRUE,
    font_size         = 11
  ) |>
  kableExtra::row_spec(fail_rows, background = "#fde8e8", bold = TRUE) |>
  kableExtra::row_spec(warn_rows, background = "#fff8e1")
```

------------------------------------------------------------------------

## Notes on the 30×30 Conserved Areas Data

The Terrestrial 30×30 Conserved Areas layer (CA Nature, 2024) integrates:

- **CPAD** (California Protected Areas Database, GreenInfo Network)
- **CCED** (California Conservation Easement Database, GreenInfo Network)
- **PAD-US** (USGS Gap Analysis Project) for GAP status coding

GAP status was assigned using an 80% areal overlap rule at the Superunit
level. Areas with less than 80% overlap with any single GAP code were resolved
at the Holdings level. Remaining uncoded areas were assigned GAP 4.

Layer citation: *California Protected Areas Database (CPAD). December 2024.
GreenInfo Network. www.calands.org*
