---
title: "BeeMarkdown"  # Document title
author: "Kap"  # Author's name
date: "2024-07-18"  # Date of creation
output: html_document  # Specifies that the output will be an HTML document
---

# This chunk ensures that required packages are installed and loaded

```{r}
# Define a list of necessary packages for this script
packages_to_install <- c("ggplot2", "rstudioapi")  
```

# Function to check if packages are installed, install missing ones, and load all required packages

```{r}
# Function to check if packages are installed, install missing ones, and load all required packages
install_and_load_packages <- function(packages) {
  # Identify packages that are missing by comparing to installed packages
  packages_needed <- packages[!(packages %in% installed.packages()[,"Package"])]  
  # If any packages are missing, install them
  if (length(packages_needed) > 0) {  
    install.packages(packages_needed, dependencies = TRUE)  # Install packages with dependencies
    print(paste("Installed packages:", paste(packages_needed, collapse = ", ")))  # Print installed packages
  } else {
    print("All required packages are already installed.")  # Message if no installation was needed
  }
  # Load each package into the session
  invisible(lapply(packages, library, character.only = TRUE))  
}

# Execute the function to check and install required packages before proceeding
install_and_load_packages(packages_to_install) 
```

# Setting up workspace: Select the folder containing "Data" and this RMD file

```{r}
#dir_path <- selectDirectory()  # Opens a dialog box to select a folder
#setwd(dir_path)  # Sets the working directory to the chosen folder
```

# Load bee dataset from the "Data" folder

```{r}
# Read the CSV file containing bee species and climate data
beeData <- read.csv("~/bees_climate/data/proccessed/BeeDataNoiNatBio_PPTCWE.csv")  

# Display the first few rows of the dataset to check its structure
head(beeData)  
```

# Graph: Species Richness vs. Average Spring Precipitation (March-October)

```{r}
ggplot(data = beeData, aes(x = X_RICHNESS_ALL_, y = PPT)) +  # Set x-axis (species richness) and y-axis (precipitation)
  geom_point(color = "black", pch = 21) +  # Add scatter plot points
  geom_smooth(method = "lm", se = FALSE, color = "maroon") +  # Add a linear regression line without shaded error bands
  labs (
    title = "Species Richness and Average Spring Precipitation",  # Main graph title
    subtitle = "Per 15 km cells across California"  # Subtitle describing the dataset coverage
  ) + 
  xlab("Species Richness (# of species per cell)") +  # Label for x-axis
  ylab("Average Precipitation (mm)")  # Label for y-axis
```

# Graph: Corrected Weighted Endemism (CWE) vs. Average Spring Precipitation (March-October)

```{r}
ggplot(data = beeData, aes(x = CWE, y = PPT)) +  # Set x-axis (CWE) and y-axis (precipitation)
  geom_point(color = "black", pch = 21) +  # Add scatter plot points
  geom_smooth(method = "lm", se = FALSE, color = "maroon")  # Add a linear regression line without shaded error bands
  labs (
    title = "Corrected Weighted Endemism and Average Spring Precipitation",  # Corrected title for clarity
    subtitle = "Per 15 km cells across California"  # Subtitle describing dataset coverage
  )
  xlab("Corrected Weighted Endemism (CWE)")  # Label for x-axis
  ylab("Average Precipitation (mm)")  # Label for y-axis
```
