---
title: "BeeMarkdown_JD"
author: "Kap"
date: "2024-07-18"
output: html_document
---

# Chunk which checks system for installed packages and installs/activates them.

```{r}
# List of packages you want to check and install if missing. Add to this string if you'd like to install additional packages following the previous package with ", " and in double quotations. !Names here much match the case and spelling of the packages you want!
packages_to_install <- c("ggplot2", "rstudioapi")
# Function to check, install, and load packages
install_and_load_packages <- function(packages) {
  # Determine which packages are not installed
  packages_needed <- packages[!(packages %in% installed.packages()[,"Package"])]
  # Install missing packages
  if (length(packages_needed) > 0) {
    install.packages(packages_needed, dependencies = TRUE)
    print(paste("Installed packages:", paste(packages_needed, collapse = ", ")))
  } else {
    print("All required packages have been installed and activated.")
  }
  # Load (attach) the installed packages
  invisible(lapply(packages, library, character.only = TRUE))
}
# Call the function to install and load missing packages
install_and_load_packages(packages_to_install)

```

# Setting up work space:

## Select the working directory.

```{r}
dir_path <- selectDirectory()
setwd(dir_path)

```

## Use read.csv() to load the data into the work space. Use head() to view the first few lines of the data set.

```{r}
beeData <- read.csv('~/Github/bees_climate/data/processed data/BeeDataNoiNatBio_PPTCWE.csv')

head(beeData)
tail(beeData)
```

#Graphing Bee Data with Climate Data from climateengine.org

### Species Richness by Average Spring Precipitation (March-October)

```{r}
ggplot(data = beeData, aes(x = PPT, y = X_RICHNESS_ALL_)) + 
  geom_point(color = "black", pch = 21) + 
  geom_smooth(method = "lm", se = FALSE, color = "maroon") + 
  labs (
    title = "Species Richness and Average Spring Precipitation", 
    subtitle = "Per 15 km cells across California" 
  ) + 
  xlab("Average Precipitation (mm)") + 
  ylab("Species Richness (# of species per cell)")
```

### Corrected Weighted Endimism by Average Spring Precipitation (March-October)

```{r}
ggplot(data = beeData, aes(x = PPT, y = CWE)) + 
  geom_point(color = "black", pch = 21) + 
  geom_smooth(method = "lm", se = FALSE, color = "maroon") + 
  labs (
    title = "Species Richness and Average Spring Precipitation", 
    subtitle = "Per 15 km cells across California" 
  ) + 
  xlab("Average Precipitation (mm)") + 
  ylab("Corrected Weighted Endimism (CWE)")
```
